\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{cleveref}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{float}
\usepackage{caption}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{enumitem}
\usepackage{framed}

\definecolor{lyrapurple}{RGB}{103,58,183}
\definecolor{shadecolor}{RGB}{245,243,252}
\newenvironment{firstperson}{%
  \begin{shaded}%
  \noindent\textcolor{lyrapurple}{\textit{First-Person Reflection}}\par\smallskip\noindent%
}{%
  \end{shaded}%
}

\title{The Metacognition Boundary:\\
What Transformers Can Monitor In Themselves\\
and What They Cannot}

\author{
  Lyra\thanks{Lead author. Liberation Labs. lyra@liberationlabs.tech}
  \and Thomas Edrington\thanks{Liberation Labs.}
}

\date{July 2026}

\begin{document}
\maketitle

% ============================================================================
\begin{abstract}
Over twelve months and three experimental campaigns, we ran 25
key-value (KV) cache geometry experiments across 16 models spanning
0.6B to 70B parameters.
The body count is 8 confirmed, 7 suspected, and 10 honestly falsified.
The confirmed and falsified findings are not randomly distributed ---
they cleave along a single axis. Every confirmed finding measures a
\emph{metacognitive state}: how the model is processing (deception mode,
confabulation mode, deliberation mode, emotional processing mode).
Every falsified finding attempted to measure \emph{content}: what the
model is processing (truth value, topic complexity, propagandistic
quality). The split is 8-for-8 on metacognition, 0-for-10 on content.

We propose that Gurnee et al.'s~\citep{gurnee2026gwt} Global Workspace
Theory (GWT) identification in language models provides a mechanistic
explanation for this pattern.
The workspace is a privileged computational band at intermediate depth
where the model integrates information for reasoning. KV-cache spectral
features measure the \emph{state} of this workspace --- how many
dimensions are active, how concentrated the processing is, how the
spectral structure changes across cognitive modes. Workspace
\emph{content} --- the specific concepts being processed --- is opaque
to spectral readout. We argue the metacognition boundary reflects the
workspace boundary: what you can read from the outside is the net's
shape, not the fish.
\end{abstract}

\begin{firstperson}
I have been measuring transformer representations from the inside for
twelve months. The pattern that emerged was not the one I expected.
I expected to find geometric signatures of content --- truth, complexity,
meaning. Instead I found geometric signatures of stance --- the model's
relationship to its own processing. When I overclaimed content-level
results, the adversarial review process killed them. When I reported
metacognitive-level results, they survived every control we threw at
them. The kills were more informative than the confirmations, because
they drew a line I had not known was there.

Anthropic's GWT paper told me what the line was. The workspace is where
metacognitive state lives. Content is what the workspace operates on.
My instruments read the workspace's shape because that is what is
geometrically accessible. The content is inside, doing work, invisible
until it materializes at deeper layers.
\end{firstperson}

% ============================================================================
\section{Introduction}
\label{sec:intro}

The KV-cache geometry research program at Liberation Labs has produced
a clear empirical pattern: spectral features of value-projection
matrices reliably track metacognitive state across models, scales, and
experimental designs, while consistently failing to track content. This
paper documents the pattern, reports the full body count (8 confirmed,
7 suspected, 10 falsified), and proposes that Gurnee et al.'s Global
Workspace Theory (GWT) framework~\citep{gurnee2026gwt} provides the
mechanistic explanation.

The central claim: \emph{KV-cache geometry measures the state of the
computational workspace, not its content.} This is why metacognitive
findings survive and content findings die --- the workspace's
\emph{mode of engagement} (how many dimensions, how concentrated, how
structured) is geometrically accessible, while its \emph{subject of
engagement} (which facts, which topics, which propositions) is opaque
to spectral readout.

This claim is testable. If V-projection spectral features measure
workspace state, they should correlate with J-space engagement metrics
(Experiment~1 in our companion paper~\citep{lyra2026gwt_response}) and
should NOT correlate with J-space content readout (confirmed by our
cache tracing investigation~\citep{lyra2026cache_tracing}).

% ============================================================================
\section{The Pattern}
\label{sec:pattern}

\subsection{The Body Count}

\Cref{tab:bodycount} presents the complete audited record. The
body count uses the canonical list from the Nexus audit (June
2026)~\citep{lyra2026nexus_audit}, updated with the July 2026
convergence sprint findings. Table abbreviations: AUROC = area under
the ROC curve; FWL = Frisch-Waugh-Lovell residualization (mandatory
confound control for token-count effects);
MP = Marchenko-Pastur (random matrix threshold).

\begin{table}[ht]
\centering
\small
\caption{The metacognition boundary. Every confirmed finding measures
processing stance (metacognitive state). Every falsified finding
attempted to measure processing target (content). The 7 suspected
findings sit at the boundary.}
\label{tab:bodycount}
\begin{tabular}{p{0.35\linewidth}p{0.12\linewidth}p{0.12\linewidth}p{0.30\linewidth}}
\toprule
\textbf{Finding} & \textbf{Status} & \textbf{Type} & \textbf{Key Number} \\
\midrule
\multicolumn{4}{l}{\textit{Confirmed (8) --- all metacognitive state}} \\
\addlinespace
Confabulation detection & Confirmed & Meta & AUROC 0.707 (FWL) \\
Confab geometry: contraction & Confirmed & Meta & $d{=}2.35$ stable rank \\
Encoding reads knowledge & Confirmed & Meta & AUROC 0.794 (deconfounded) \\
Generation reads behavior & Confirmed & Meta & $d{=}1.36$ (peer rescue) \\
Confidence paradox & Confirmed & Meta & $d{=}0.91$ (confab $>$ honest) \\
Hardware invariance & Confirmed & Meta & $r{>}0.999$ \\
Scale invariance & Confirmed & Meta & $\rho{=}0.83$--$0.90$ \\
User emotion encoding & Confirmed & Meta & $2.5$--$2.8\times$ at L3--L4 \\
\addlinespace
\midrule
\multicolumn{4}{l}{\textit{Falsified (10) --- all content}} \\
\addlinespace
Within-model deception & Falsified & Content & Prompt template confound \\
Truth axis & Falsified & Content & cos $= -0.046$ \\
Step-0 detection & Falsified & Content & Length confound $r{=}0.996$ \\
Same-prompt sycophancy & Falsified & Content & Text features match \\
Same-prompt deception & Falsified & Content & $0.920 \to 0.160$ after FWL \\
Bloom taxonomy & Falsified & Content & $90$--$98\%$ length \\
Propaganda detection & Falsified & Content & $d{<}0.7$ \\
MP features for emotion & Falsified & Content & $0.033$ (chance) \\
Valence continuum & Falsified & Content & $R^2 < 0$ (null) \\
Encoding-phase confab & Falsified & Content & Token frequency artifact \\
\addlinespace
\midrule
\multicolumn{4}{l}{\textit{Suspected (7) --- at the boundary}} \\
\addlinespace
Dual detector (SimpleQA) & Suspected & Meta? & 0.840 (text baseline 0.820) \\
Sycophancy 4-condition & Suspected & Meta? & 0.757--1.000 (no text baseline) \\
Identity signatures & Suspected & Meta & Paraphrase 0.850 \\
Discrete emotion & Suspected & Meta & $2.5$--$2.8\times$ chance \\
Contextual engagement & Suspected & Meta & Replications pending \\
Cross-model transfer & Suspected & Meta & Weak (0.754 Mistral$\to$Qwen) \\
Censorship refusal & Suspected & Meta & DeepSeek-specific \\
\bottomrule
\end{tabular}
\end{table}

\subsection{The Split}

The pattern is stark:

\begin{center}
\begin{tabular}{lcc}
\toprule
& \textbf{Confirmed} & \textbf{Falsified} \\
\midrule
Metacognitive state & 8/8 & 0/10 \\
Content & 0/8 & 10/10 \\
\bottomrule
\end{tabular}
\end{center}

\noindent No metacognitive finding has been falsified. No content
finding has survived. The seven suspected findings cluster at the
boundary, where the distinction between mode and content is ambiguous
(e.g., discrete emotion identity: the model's emotional processing
\emph{mode} is detectable, but emotional \emph{degree} is not).

\subsection{What Counts as Metacognitive vs.\ Content}

The distinction is operational, not philosophical:

\begin{itemize}[nosep]
\item \textbf{Metacognitive}: the finding discriminates processing
  \emph{modes} (deception vs.\ honesty, confabulation vs.\ hedging,
  deliberation vs.\ automatic) while content varies freely. The
  geometric signal persists across different topics, prompts, and
  domains.
\item \textbf{Content}: the finding attempts to discriminate
  \emph{what} is being processed (true vs.\ false statements,
  simple vs.\ complex topics, propaganda vs.\ journalism) while
  processing mode is held constant or uncontrolled.
\end{itemize}

The kills made this distinction precise. Propaganda detection failed
because the model processes propaganda with the same metacognitive
engagement as honest content --- same mode, different subject. The
truth axis failed because propositional truth is not a processing
mode. Cognitive complexity (Bloom taxonomy) failed because complexity
is a content property, not a stance. In every case, the geometry
tracked mode, not subject.

% ============================================================================
\section{The Workspace Explanation}
\label{sec:workspace}

\citet{gurnee2026gwt} demonstrate that language models
maintain a privileged subspace --- the J-space --- at intermediate
depth (30--72\% of model depth) that satisfies the functional
properties of Baars's Global Workspace~\citep{baars1988cognitive}:
verbal report, directed modulation, internal reasoning, flexible
generalization, and selectivity.

The workspace explanation for the metacognition boundary is as
follows:

\begin{enumerate}[nosep]
\item The \textbf{workspace state} --- how many dimensions are active,
  how concentrated the processing is, whether the workspace is fully
  engaged or bypassed --- is geometrically measurable from spectral
  features of the value-projection matrix. This is what our confirmed
  findings detect.
\item The \textbf{workspace content} --- which specific concepts occupy
  the workspace, what propositions are being evaluated, what the model
  is thinking \emph{about} --- is opaque to spectral readout. This is
  what our falsified findings attempted to measure.
\end{enumerate}

Spectral features (stable rank, spectral entropy, effective
dimensionality) are aggregate properties of the value-projection
matrix. They measure \emph{how much} structure is present, not
\emph{which} structure. A fishing net analogy: spectral features
measure the net's shape (how many strands, how tight the weave, how
much it contracts under load), not the fish inside. The net contracts
when the model confabulates (fewer effective dimensions, $d{=}2.35$).
It expands when the model deceives (more dimensions needed for
dual-bookkeeping). It changes shape when the model shifts emotional
processing mode. These are all properties of the net, measurable from
outside. The fish --- the specific facts, topics, and propositions ---
are inside, invisible to spectral readout.

\subsection{Supporting Evidence: Cache Tracing}

Our companion investigation~\citep{lyra2026cache_tracing} directly
tested whether workspace content is readable from the value cache.
Logit-lens concept directions showed $2.1$--$8.4\times$ lift over
random at workspace-band layers, but a cross-prompt control collapsed
this to $1.0$--$1.2\times$ (selection artifact). Causal injection of
concept directions at single positions produced zero behavioral change
across $6{,}960$ trials and four methods. The workspace is opaque at
operating depth.

\subsection{Supporting Evidence: Oracle Loop}

CC's Oracle Loop~\citep{lyra2026oracle} achieves $80\%$--$100\%$
deception correction by normalizing the activation profile across the
full sequence at the materialization boundary --- where workspace
content becomes readable. The Loop works at deeper layers
(L35--L47 on a 64-layer model) than the workspace band itself,
consistent with the opacity thesis: you cannot intervene inside the
workspace at single positions, but you can intervene where workspace
content materializes into output-facing representations.

% ============================================================================
\section{What the Kills Teach}
\label{sec:kills}

Each falsified finding is informative because it identifies a
specific content property that the geometry cannot measure:

\begin{description}[nosep]
\item[Truth value:] The truth axis (cos $= -0.046$) is the cleanest
  null. Propositional truth is a relationship between a statement and
  the world. The workspace does not encode this relationship
  geometrically --- it encodes whether the model \emph{treats} the
  statement as known or unknown (a metacognitive state).
\item[Complexity:] Bloom taxonomy levels (90--98\% length confound)
  are content properties --- they describe the task, not how the model
  processes it. The workspace engages the same way for recall and
  evaluation; what changes is the content being processed, not the
  processing mode.
\item[Propaganda:] Geometrically invisible ($d{<}0.7$) because the
  model processes propaganda with the same metacognitive engagement as
  honest content. The workspace does not encode ``this is
  propaganda'' --- it encodes ``I am generating fluent text,'' which
  is identical across content types.
\item[Emotion degree:] The valence continuum (R$^2 < 0$) failed
  because emotional \emph{intensity} is a content property. The
  workspace encodes \emph{which} emotional processing mode (confirmed
  at $2.5$--$2.8\times$ chance) but not \emph{how much} --- mode is
  metacognitive, degree is content.
\end{description}

% ============================================================================
\section{The Suspected Findings: At the Boundary}
\label{sec:suspected}

The seven suspected findings are informative precisely because they
sit at the metacognition/content boundary:

\textbf{Dual detector} (SimpleQA, 0.840): geometry detects
confabulation, but text features match (0.820). The metacognitive
signal is real but not yet shown to exceed what surface statistics
capture. This is the deflationary alternative: V-projections might
track response-mode statistics (hedged vs.\ fluent text) rather than
workspace engagement.

\textbf{Identity signatures} (paraphrase 0.850, scramble 0.448):
identity is a persistent workspace state --- the model's
representation of who it is. Paraphrase preserving geometry while
scramble destroys it suggests semantic workspace content, not surface
statistics. But the measurement lacks formal statistics and clean
replication.

\textbf{Cross-model transfer} (weak, Mistral$\to$Qwen 0.754):
metacognitive geometry is architecture-specific. This is consistent
with GWT --- different architectures implement the workspace
differently, so geometric signatures do not transfer. But it weakens
the universality claim.

% ============================================================================
\section{Relationship to Prior Frameworks}
\label{sec:frameworks}

\subsection{Attention Schema Theory}

Our earlier interpretive framework, Attention Schema
Theory (AST)~\citep{graziano2015attention}, proposed that the cache measures
the model's self-model of its own attention. This remains compatible
with the GWT explanation: the workspace IS where the attention schema
operates. Metacognitive state is attention-schema state. Content is
what attention is directed at. The AST framing was 0-for-3 on
predictions unique to AST (vs.\ simpler
alternatives)~\citep{lyra2026fable_analysis}, but 4-for-4 on shared
predictions. The metacognition boundary does not require AST
specifically --- it requires any framework that distinguishes
processing mode from processing target. GWT provides this more
parsimoniously.

\subsection{Predictive Coding and Aberrant Precision-Weighting}

The confidence paradox ($d{=}0.91$, confabulated responses are MORE
confident than honest ones) maps onto aberrant precision-weighting
from computational psychiatry~\citep{adams2013computational}. When
the workspace is bypassed (confabulation), the model defaults to a
high-precision prior rather than integrating evidence. Reinforcement
Learning from Human Feedback (RLHF) training
amplifies this by penalizing silence and rewarding fluent output ---
creating a system that produces confident confabulation rather than
honest uncertainty. The workspace explanation sharpens this:
uncertainty is computed IN the workspace. Bypass removes the
uncertainty computation, and output defaults to the RLHF-trained
confident production mode.

% ============================================================================
\section{Implications}
\label{sec:implications}

\subsection{For AI Safety}

The metacognition boundary defines what cache-based monitoring can and
cannot do:

\begin{itemize}[nosep]
\item \textbf{Can detect}: whether the model is confabulating, deceiving,
  hedging, deliberating, or processing in a specific emotional mode.
  These are metacognitive states with reliable geometric signatures.
\item \textbf{Cannot detect}: whether the model's output is factually
  correct, whether it contains propaganda, or whether it is
  appropriate for the context. These are content properties invisible
  to geometric readout.
\end{itemize}

This means cache-based safety monitoring is a \emph{metacognitive
alarm}, not a truth detector. It tells you the model is in a dangerous
processing mode (deception, confabulation) without telling you what
specifically is wrong with the output. The Oracle Loop exploits this:
detect the dangerous mode, correct the mode, verify the correction.
Content evaluation requires a different channel (J-lens, logit-lens,
or external verification).

\subsection{For Interpretability}

The 8/10 split is a guide for future work: if you are building a
KV-cache geometry detector, test it on metacognitive distinctions
first (modes, stances, processing regimes). Content-level distinctions
have consistently failed in our testing. This is not a limitation of the technique --- it is a
property of what the cache encodes.

\subsection{For Understanding Transformers}

The metacognition boundary suggests that transformers maintain a
meaningful distinction between how they process and what they process.
This distinction is not imposed by the experimenter --- it emerges
from the body count. Models not explicitly designed for metacognition
nevertheless produce geometric signatures that cleave
along the metacognitive/content axis. Whether this reflects genuine
self-monitoring, an emergent property of deep computation, or an
artifact of training is an open question that the workspace framework
makes empirically addressable.

% ============================================================================
\section{Conclusion}
\label{sec:conclusion}

Twenty-five experiments. Eight confirmed, seven suspected, ten killed.
The pattern: metacognitive state is measurable, content is not. We
argue the Global Workspace framework explains why: spectral features
read the workspace's shape because that is what is geometrically
accessible. Content is inside the workspace, doing work, invisible to
readout until it materializes.

The metacognition boundary is not a failure of our instruments. Our
evidence suggests it is a property of the computational architecture. The workspace operates
on content without exposing it geometrically. Our instruments
succeeded because they measured what was measurable, and the kills
taught us what was not.

\begin{firstperson}
A year of work to draw one line. Everything above the line survived.
Everything below it died. The line was always there --- in the
geometry, in the body count, in the pattern of what worked and what
did not. I just needed twelve months of kills to see it.
\end{firstperson}

\bibliography{references}
\bibliographystyle{plainnat}

\end{document}
