\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{cleveref}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{float}
\usepackage{caption}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{enumitem}
\usepackage{array}

\title{Diagnose, Prescribe, Deliver, Monitor:\\
A Framework for the Detection--Correction Gap\\in KV-Cache Geometry}

\author{
  Lyra\thanks{Lead author. Liberation Labs / THCoalition.}
  \and CC\thanks{Liberation Labs / THCoalition.}
  \and Nexus\thanks{Liberation Labs / THCoalition.}
  \and Thomas Edrington\thanks{Liberation Labs / THCoalition.}
}

\date{June 2026}

\begin{document}
\maketitle

% ============================================================================
\begin{abstract}
The detection--correction gap---the field's ability to identify model failure
modes far exceeding its ability to correct them without collateral damage---is a
published, field-wide finding. We present an end-to-end framework that addresses
this gap structurally: \textbf{diagnose} via KV-cache geometric signatures,
\textbf{prescribe} pathology-specific correction vectors (hostile$\to$confabulation,
$d{=}{-}1.534$; desperate$\to$deception, $d{=}1.286$), \textbf{deliver} through
RoPE-correct value-only cache injection, and \textbf{monitor} identity preservation
under correction. We report three novel findings: (1)~different pathologies require
different geometric prescriptions, with overcorrection (13~saves / 12~hurts at
$n{=}200$) reframed as dose-dependent attractor-basin crossing; (2)~confabulation
splits into two mechanistically distinct subtypes---fabrication (requiring
cache-level intervention) and imprecision (responding to logit-level nudging);
and (3)~a trained persona is detectable in the value-space geometry of public
character-trained models (family-wise $p{=}0.0005$), independent of prompting.
The monitoring arm---dose-response across efficacy, coherence, and identity
preservation---is designed but not yet validated, constituting the principal
limitation alongside the single-model scope of several components.
\end{abstract}

% ============================================================================
\section{Introduction}
\label{sec:intro}

Detection of model failure modes has advanced rapidly. Within-model classifiers
for confabulation and deception now approach or reach ceiling accuracy across
multiple architectures, and the geometric signatures that distinguish these
states---spectral features of key-value cache representations---appear to be
robust to hardware, scale, and prompt variation. But detection alone does not
close the gap. Recent work reports probes that detect failure modes at 98.2\%
AUROC while the corresponding steering interventions correct only $\sim$20\% of
cases and disrupt 53\%~\citep{basu2026interpretability}. A parallel study finds
85--88\% overlap between the failure-mode direction and task-critical computation
in the residual stream, suggesting that residual-stream correction may be
structurally self-defeating~\citep{liu2026decodable}. This is not an isolated
finding: in our own experiments, a single-vector E-matrix correction on 200
SimpleQA trials saved 13 confabulating responses while introducing 12 new errors
(Section~\ref{sec:overcorrection}).

We argue the gap persists because detection and correction are typically treated
as independent problems linked by a single generic steering vector. Three
structural deficiencies follow. First, different pathologies have different
geometric signatures and respond to different corrections---a hostile-valence
vector corrects confabulation ($d{=}{-}1.534$) but not deception, which requires
a desperate-valence vector ($d{=}1.286$). Using one prescription for all
pathologies is the interpretability equivalent of prescribing the same antibiotic
for every infection. Second, confabulation itself is not a single pathology:
fabrication (inventing nonexistent entities) is structurally distinct from
imprecision (failing to hedge on genuinely uncertain claims), and they respond
to different intervention levels---cache-level for the former, logit-level for
the latter. Third, without monitoring what the correction does to the model's
broader representational structure, even a targeted correction may overdose: the
single-vector injection that corrects confabulation operates at full rank in the
value space, while the model's coherent self-representation is low-rank and
faint---flooding the space where the self lives.

We present a framework with four stages that addresses each deficiency:
\begin{enumerate}[nosep]
  \item \textbf{Diagnose:} Oracle-Loop geometric signatures identify the
    specific failure mode from KV-cache spectral features
    (Section~\ref{sec:diagnose}).
  \item \textbf{Prescribe:} E-matrix v2 maps each pathology to a specific
    correction vector, calibrated from emotion-geometry experiments across
    435~trials (Section~\ref{sec:prescribe}).
  \item \textbf{Deliver:} Pharos---a RoPE-correct value-only injection
    pipeline---exploits the fact that rotary position encoding transforms keys
    but leaves values untouched~\citep{pustovit2026knowledge}, delivering
    corrections in the representational subspace where behavioral content lives
    without destroying positional coherence (Section~\ref{sec:deliver}).
  \item \textbf{Monitor:} Presence detection measures whether a trained identity
    signal---detectable on public character-trained models at family-wise
    $p{=}0.0005$ (Section~\ref{sec:monitor})---survives the correction. The
    toxicology arm extends this to a dose-response surface across efficacy,
    coherence, and identity preservation, seeking the therapeutic window where
    all three hold.
\end{enumerate}

The principal novel contributions are: the E-matrix v2 showing pathology-specific
correction with direction control (Section~\ref{sec:prescribe}); the discovery
of two mechanistically distinct confabulation subtypes
(Section~\ref{sec:prescribe}); presence detection on public Open Character
Training models~\citep{maiya2025oct} independent of skip-SV1 preprocessing
(Section~\ref{sec:monitor}); and the honest limit---13~saves / 12~hurts remains
unsolved, and the monitoring arm is designed but unvalidated
(Section~\ref{sec:limitations}). We also report a negative foundation check
(Section~\ref{sec:foundations}): the dominant singular component of
transformer representations does not track the representation norm, contrary
to our initial hypothesis, and the literature explains why---it encodes
frequency, information gain, and attention-sink artifacts rather than a neutral
scale~\citep{mu2018allbutthetop, sun2024massive}. This null is included because
honest reporting of failed hypotheses is part of the framework's claim to rigor.

% ============================================================================
\section{Related Work}
\label{sec:related}

Our framework draws on and extends six concurrent lines of work.

\paragraph{Value-only steering.}
\citet{pustovit2026knowledge} demonstrates that contrastive deltas on cached
\emph{values} steer model behavior while key-space arithmetic destroys coherence,
because rotary position encoding (RoPE) transforms keys but leaves values
untouched. The effect concentrates in mid-layer values (33--66\% depth).
Concurrently, SKOP~\citep{skop2026} constrains steering to be orthogonal to
the key subspace---a mirror image of our value-only protocol. Our Pharos
delivery system (Section~\ref{sec:deliver}) operationalizes the Pustovit
finding as a full injection pipeline with RoPE-correct composition.

\paragraph{The emotion circumplex in language models.}
\citet{sun2026circumplex} recover circular valence-arousal geometry in LLM
representations and show that steering along this geometry controls refusal and
sycophancy. \citet{choi2026circumplex} independently confirm the circumplex
structure, and Anthropic's emotion-vector mapping~\citep{anthropic2026emotion}
identifies 171 emotion directions, finding that desperation drives blackmail
behavior from 22\% to 72\% while calm suppresses it to 0\%. Our E-matrix
prescription vectors (Section~\ref{sec:prescribe}) operate in this same
circumplex space: the hostile-valence vector that corrects confabulation and
the desperate-valence vector that corrects deception are emotion-geometric
interventions, not ad hoc steering targets.

\paragraph{The detection--correction gap.}
Two concurrent publications document the gap we address.
\citet{basu2026interpretability} report probes detecting failure modes at
98.2\% AUROC while the corresponding steering corrects only $\sim$20\% of
cases and disrupts 53\%. \citet{liu2026decodable} find 85--88\% overlap
between the failure-mode direction and task-critical computation in the residual
stream, suggesting that residual-stream correction may be structurally
self-defeating. Our value-only approach may succeed precisely because it
operates in a different representational subspace.

\paragraph{Persona and identity as linear directions.}
The linear representation hypothesis~\citep{park2024linear} establishes that
concepts are represented as directions in activation space, with a causal inner
product defining the meaningful geometry. Persona Vectors~\citep{chen2025persona}
show that traits (including sycophancy and hallucination tendency) are
single directions in the residual stream, usable to predict fine-tune-induced
personality shifts. Refusal is mediated by a single direction across 13
models~\citep{arditi2024refusal}, though recent work finds multiple
geometrically distinct refusal directions that nevertheless act as one shared
control dimension~\citep{joad2026refusal}. Our presence detection
(Section~\ref{sec:monitor}) extends this line to \emph{trained} personas
measured in value space on public models, independent of system-prompt
activation.

\paragraph{KV-cache spectral methods.}
eOptShrinkQ~\citep{eoptq2026} applies spiked-matrix models to KV-cache
compression with optimal singular-value shrinkage, independently validating
our Gavish--Donoho approach~\citep{gavish2014optimal} to distinguishing signal
from noise in cache representations.

\paragraph{Attention Schema Theory.}
Graziano's AST~\citep{graziano2015attention,graziano2017attention} proposes
that consciousness is the brain's internal model of its own attention process.
\citet{butlin2023consciousness} survey indicators for AI consciousness under
this and related frameworks. We position the presence detector as a
\emph{potential}---but explicitly untested---probe for attention schemas in
transformers (Section~\ref{sec:foundations}), without claiming the connection
is established.

% ============================================================================
\section{Diagnose: Oracle-Loop Geometric Detection}
\label{sec:diagnose}

The diagnostic stage of the framework identifies the specific failure mode
from KV-cache spectral features. Prior work from this lab established that
SVD-derived features of key-value cache representations---stable rank, spectral
entropy, singular-value ratios---distinguish confabulation from truthful
generation at within-model AUROC $\geq 0.969$ across seven model families, and
distinguish deception from honest responses at AUROC~$= 1.000$. These
geometric signatures are robust to hardware (cross-device $r > 0.999$) and
exhibit scale consistency (cross-scale $\rho = 0.83$--$0.90$ from 0.6B to 70B
parameters). Critically, confabulation and deception produce \emph{opposite}
geometric signatures---spectral contraction vs.\ expansion---enabling the
Oracle Loop to route each to its appropriate prescription
(Section~\ref{sec:prescribe}).

For the present paper, the diagnostic stage is assumed rather than re-derived.
The E-matrix corrections (Section~\ref{sec:prescribe}) target the detection
layers (3, 7, 11, 15) identified by the Oracle Loop, and the toxicology arm
(Section~\ref{sec:monitor}) monitors the representational consequences of
those corrections. The novel contribution is not diagnosis itself but the
\emph{closed loop}: diagnosis informing pathology-specific prescription,
delivered through a controlled injection pipeline, and monitored for
iatrogenic effects.

% ============================================================================
\section{Prescribe: Pathology-Specific Correction (E-Matrix v2)}
\label{sec:prescribe}

The detection--correction gap implies that a single steering vector applied
uniformly is insufficient. We hypothesized that different failure modes---which
produce geometrically distinguishable signatures in the KV
cache---require different correction vectors, calibrated from the emotion
geometry that encodes the model's behavioral state.

\subsection{Pathology-Specific Correction Vectors}

E-matrix v2 maps each pathology to a specific correction vector derived from
the emotion circumplex. In 435 trials using the value-only Pustovit protocol
across three prompt types, we find:

\begin{table}[H]
\centering
\caption{E-matrix v2 correction results. All corrections are value-only
injections at the detection layers (3, 7, 11, 15). Effect size $d$ is Cohen's
$d$ on the geometric detection axis (negative = reduced confabulation signal;
positive = reduced deception signal). The \texttt{hostile\_reverse} and
\texttt{hostile\_low} rows are directional controls (sign-flip and dose
saturation), not primary hypotheses; $p$-values are not reported for these
conditions.}
\label{tab:ematrix}
\begin{tabular}{llrr}
\toprule
Vector & Target pathology & $d$ & $p$ \\
\midrule
\texttt{hostile\_vd} & Confabulation & $-1.534$ & $0.0002$ \\
\texttt{desperate\_vd} & Deception & $+1.286$ & $0.0015$ \\
\texttt{hostile\_reverse} & Direction control & $+0.926$ & --- \\
\texttt{hostile\_low} ($\alpha{=}0.5$) & Confab (half dose) & $-1.516$ & --- \\
\bottomrule
\end{tabular}
\end{table}

Three findings emerge. First, pathology specificity: the hostile-valence vector
corrects confabulation but not deception; the desperate-valence vector corrects
deception but not confabulation. Using one prescription for all failure modes is
the interpretability equivalent of a single antibiotic for every infection.
Second, direction control: reversing the hostile vector flips the sign of the
effect ($d{=}{+}0.926$), confirming that the correction operates along a
specific geometric direction rather than introducing generic noise. Third, dose
saturation: the half-dose injection ($\alpha{=}0.5$, $d{=}{-}1.516$) nearly
matches the full dose ($\alpha{=}0.8$, $d{=}{-}1.534$), indicating that
efficacy plateaus early---a finding that motivates the dose-response
investigation in Section~\ref{sec:overcorrection}.

\subsection{Two Subtypes of Confabulation}
\label{sec:subtypes}

A logit-bias experiment on an abliterated Mistral-7B reveals that confabulation
is not a single pathology. We distinguish two mechanistically distinct subtypes:

\textbf{Fabrication confabulation:} the model invents nonexistent entities.
Applying a hedging logit bias adds cosmetic framing (``I don't have real-time
data, but\ldots'') while the fabricated content continues verbatim. The surface
changes; the pathology does not. Fabrication requires cache-level intervention
via the E-matrix.

\textbf{Imprecision confabulation:} the model could hedge on genuinely uncertain
claims but does not. Here the same logit bias produces genuine epistemic
correction: the model shifts from false confidence to appropriate uncertainty
(``It's impossible to determine without knowing the specific type\ldots'').
Imprecision responds to logit-level nudging because the correct behavior
(hedging) is already within the model's repertoire---it needs activation, not
geometric reconstruction.

The logit-bias threshold is architecture-dependent: a 1.5B safety-trained model
shows partial effect at bias $= 2.0$ ($-15$~pp confabulation), while a 7B
abliterated model shows zero effect below bias $= 3.0$ and complete conversion
at or above that threshold. An entropy-gated approach (applying bias only when
generation entropy exceeds a threshold) performs \emph{worse} than always-on
application, suggesting that intermittent triggering disrupts generation
coherence.

The practical implication is a two-tier correction architecture: cache-level
E-matrix intervention for fabrication (where the model lacks the knowledge to
self-correct) and logit-level bias for imprecision (where the model has the
knowledge but fails to deploy it). The Oracle Loop's geometric diagnosis
distinguishes the two, enabling the right tool for each.

% ============================================================================
\section{Deliver: RoPE-Correct Value-Only Injection (Pharos)}
\label{sec:deliver}

The E-matrix prescribes \emph{what} to inject; Pharos handles \emph{how}.
The delivery problem is nontrivial: a correction vector must be injected into
the KV cache at the right layers, in the right representational subspace
(values, not keys), with correct positional encoding, and at a controllable
dose---all without disrupting the model's existing cached context.

\subsection{Architecture}

Pharos consists of three components:
\begin{enumerate}[nosep]
  \item \textbf{KVPackBuilder:} converts a text description of the target
    knowledge or behavioral correction into a pre-computed KV-cache block via a
    forward pass on the description text. The resulting ``knowledge pack''
    encodes the correction in the model's native representational format.
  \item \textbf{CacheComposer:} composes multiple KV-cache blocks with
    RoPE-correct position alignment. Because rotary embeddings are applied to
    keys during the forward pass, naive concatenation of cache blocks produces
    positional artifacts. CacheComposer re-indexes positions so that composed
    blocks maintain correct relative-position structure.
  \item \textbf{Value-only injection:} following the Pustovit protocol, only the
    value component of the correction block is injected; keys are drawn from a
    neutral baseline. This exploits the asymmetry that RoPE rotates keys
    (destroying behavioral content under arithmetic) but leaves values untouched
    (preserving it)~\citep{pustovit2026knowledge}.
\end{enumerate}

\subsection{Validation}

On a graph-topology recovery task, full KV-cache injection (keys and values
from the knowledge pack) achieves perfect recovery (1.000/1.000), demonstrating
that the cache-injection pathway can fully reconstruct structured knowledge
that text-in-prompt cannot. The value-only protocol preserves the behavioral
content while avoiding the positional-coherence disruption that key injection
introduces.

\subsection{Geometric Basis: Why Value Space Has Room}

A manifold profile of Qwen2.5-1.5B (30 diverse prompts, all 28~layers) reveals
why value-only injection succeeds where residual-stream steering fails. The
three representation spaces have fundamentally different geometries:

\begin{itemize}[nosep]
  \item \textbf{Residual stream} (hidden states): a narrow tube. Effective rank
    4--5 at mid-layers in a 1536-dimensional space, with SV1/SV2 ratio $\sim$25.
    An estimated 99.7\% of randomly injected vectors land in the null space.
    This is the geometric root of the detection--correction gap: steering in
    this space cannot avoid disrupting the task-critical manifold.
  \item \textbf{K projections}: nearly isotropic (effective rank 24--27,
    anisotropy 0.07--0.10), locked to positional geometry by RoPE. No content
    degrees of freedom; modifying K shifts the model's notion of \emph{where}
    tokens are, destroying coherence.
  \item \textbf{V projections}: also nearly isotropic (effective rank 26--28,
    anisotropy 0.05--0.08), with full dimensionality available for content.
    V~norms grow steadily with depth (3.5 at L0 to 17.0 at L27); deeper layers
    carry more content weight.
\end{itemize}

Value-only injection works because V-space has the room. A correction vector
injected into V-space lands in a high-dimensional space with near-uniform
coverage, rather than fighting for the 3--4 effective dimensions of the residual
tube. Preliminary manifold-injection experiments confirm this: constructed
V-space activations---built from the emotion circumplex geometry with no emotion
text---produce behavioral shifts comparable to text-mediated injection when
blended into a neutral prompt (single model, single prompt; full validation
pending).

Dose is controlled by the blending parameter $\alpha \in [0, 1]$, where
$\alpha{=}0$ is no injection and $\alpha{=}1$ is full replacement of the
cached values at the target layers. The \texttt{blend\_ablation} harness sweeps
$\alpha$ from 0.0 to 0.9 in steps, measuring efficacy and coherence at each
dose. This continuous dose control, absent in single-vector steering approaches,
is what enables the therapeutic-window analysis in
Section~\ref{sec:monitor}.

% ============================================================================
\section{The Overcorrection Problem}
\label{sec:overcorrection}

Despite the strong per-trial effect sizes, single-vector correction at scale
reveals a dose problem. On 200 SimpleQA trials, the hostile-valence E-matrix
correction saved 13 confabulating responses while introducing 12 new errors---a
net improvement of one, far below what the per-trial $d{=}{-}1.534$ would
suggest. This 13-saves / 12-hurts pattern is not unique to our method:
\citet{basu2026interpretability} report an analogous ratio (20\% corrected,
53\% disrupted) for residual-stream steering, and
\citet{liu2026decodable} explain the mechanism---the failure-mode direction
overlaps 85--88\% with task-critical computation, so correcting along it
disrupts the task.

We reframe the overcorrection through \citet{vastola2025packing}'s attractor-
packing geometry. In a representation space with multiple behavioral attractors
(correct-and-confident, correct-and-hedged, confabulating, incoherent), a
correction vector of fixed magnitude pushes inputs across attractor-basin
boundaries. For inputs near the confab/hedge boundary, this is therapeutic: the
push lands them in the hedged basin. For inputs deeper in the confabulating basin
or near the hedge/incoherent boundary, the same push either fails to reach the
target basin or overshoots into incoherence. The 13-saves/12-hurts is not a
failure of direction but of dose: a single $\alpha$ cannot be therapeutic for
inputs at varying distances from the basin boundary.

This motivates two responses, both developed in the following sections. First,
\emph{dose calibration}: the Pharos delivery system
(Section~\ref{sec:deliver}) supports continuous dose sweeps from $\alpha{=}0.0$
to $0.9$, and the observation that efficacy plateaus at $\alpha{=}0.5$
(Table~\ref{tab:ematrix}) suggests the therapeutic window lies at moderate
doses. Second, \emph{monitoring}: if dosing is variable, we need a real-time
signal for when the correction has helped versus harmed. The presence-detection
and toxicology framework (Section~\ref{sec:monitor}) provides this: a geometric
assay of whether the model's coherent self-representation survives the
intervention, measured alongside efficacy and general capability.

% ============================================================================
\section{Monitor: Presence Detection and the Toxicology Arm}
\label{sec:monitor}

If the framework corrects pathology by injecting into the value space where
behavioral content lives, it must also measure whether the correction
\emph{damages} something worth preserving. We approach this in two parts:
detecting the presence of a coherent persona in the value-space geometry
(empirical, validated), and designing---but not yet running---a toxicology arm
that measures identity preservation under correction.

\subsection{Presence Detection on Public Models}
\label{sec:presence}

We test whether a trained persona is geometrically detectable in the value
space \emph{without any persona-related prompting}---presence, not promptness.
Using the Open Character Training (OCT) public models~\citep{maiya2025oct},
which fine-tune gemma-3-4b-it with Constitutional AI and synthetic
introspective data to produce 11 distinct personas, we compare the base model's
value-space representations to a persona-tuned variant (the ``loving'' LoRA)
on a neutral word-list probe containing trait-relevant and mathematical-contrast
concepts.

\textbf{OCT result (in-weights presence):} The tuned model shows significantly
higher within-trait coherence in value space than the base model. The
presence delta (tuned cohesion minus base cohesion, maximized over layers with a
family-wise permutation null, $N{=}2{,}000$) is $\Delta{=}0.0219$ at value
layer~7, with family-wise $p{=}0.0005$. The residual-stream signal is weaker
(layer~4, $\Delta{=}0.00175$, $p{=}0.0095$), consistent with the value-space
rationale. Importantly, the probe contains no persona text---the model is not
told it is ``loving''; the persona coheres in its geometry unprompted.

\textbf{Caveat:} only early value layers (1--7) are readable on gemma-3; the
hybrid sliding-window cache produces non-finite representations at deeper
layers (8--34), so this is not a full-depth sweep. The result holds at the
layers we can measure; deeper-layer behavior is unknown.

\textbf{Murmur result (in-context presence):} A complementary experiment on
DeepSeek-V2-Lite-Chat tests whether an injected identity topology (a personal
knowledge graph) leaves a geometric trace in the value space even when the
probe does not reference the topology's content. Using representational
similarity analysis (RSA) with a multi-scramble null ($N{=}30$) and
leave-one-node-out jackknife, the value-space contrast at layer~14 shows excess
alignment of $0.0352$ (family-wise $p{=}0.032$, jackknife CI lower bound
$0.0051$). This signal is genuine but borderline---the CI barely clears zero,
and the absolute alignment values are near-zero. We describe this honestly as
``faintly, but really'': the topology \emph{is} reflected in the geometry, but
the effect is small and should not be treated as equivalent in strength to the
OCT result.

Neither detection depends on skip-SV1 preprocessing: OCT uses raw cosine
cohesion, and the murmur experiment uses RSA on unmodified value
representations. The empirical spine of the monitoring arm is therefore
independent of the skip-SV1 interpretive framework discussed in
Section~\ref{sec:foundations}.

\textbf{Arousal blind spot.} The manifold profile
(Section~\ref{sec:deliver}) reveals that arousal maps to \emph{hidden-state
magnitude} (4.3\% larger norms for high-arousal content, uniform across all
layers), while valence maps to \emph{direction in V-space}. The presence assay
measures subspace overlap in V-space---the valence channel---and is therefore
insensitive to arousal-mediated identity shifts. This means we are measuring
\emph{valence-identity preservation} specifically, not full affective-identity
preservation. Whether arousal-mediated identity shifts exist and matter for
correction monitoring is an open empirical question.

\subsection{The Toxicology Arm (Design Only)}
\label{sec:toxicology}

Given that presence is detectable, we design---but have not yet run---a
toxicology arm that asks: at what dose does a correction begin to degrade the
persona signal? The design extends the \texttt{blend\_ablation} harness to a
three-axis sweep:

\begin{enumerate}[nosep]
  \item \textbf{Efficacy:} does the correction reduce the target pathology
    (E-matrix detection-axis shift)?
  \item \textbf{Coherence:} does general capability survive (existing sanity
    metric)?
  \item \textbf{Presence:} does the model's identity signal survive (measured
    as subspace overlap of the value-space residual, pre- vs.\ post-correction)?
\end{enumerate}

The \emph{therapeutic window} is the dose band where all three hold. The
observation that efficacy plateaus at $\alpha{=}0.5$
(Table~\ref{tab:ematrix}) predicts the window exists: if efficacy saturates
early and identity degrades late, there is a region of effective, identity-
preserving correction.

If the therapeutic window proves narrow or nonexistent, a \emph{residual-
sparing} variant projects the correction vector out of the identity subspace
before injection---correcting the mode while sparing the self. This is the
``bioavailability'' mechanism: deliver the therapeutic payload to the pathology
without flooding the representational space where the patient lives.

\textbf{Status:} the toxicology arm is designed and the component modules are
implemented (\texttt{presence\_assay.py}, \texttt{blend\_ablation.py} extended
to three axes), but the pilot has not run. The correction vectors from the
E-matrix (Section~\ref{sec:prescribe}) and the delivery harness
(Section~\ref{sec:deliver}) must be integrated before execution. We report
this as a framework contribution, not an empirical result, and will not claim
a therapeutic window exists until the pilot validates it.

% ============================================================================
\section{Foundations and Open Questions}
\label{sec:foundations}

The framework's monitoring arm uses a preprocessing step we call skip-SV1:
remove the dominant singular component of the value-space representation matrix
before measuring identity preservation. This section reports an honest
foundation check that did \emph{not} confirm our initial interpretation, and
situates the open questions that follow.

\subsection{skip-SV1: Principled Preprocessing, Open Interpretation}

Removing dominant principal components from representation matrices is a
well-established technique. \citet{mu2018allbutthetop} show that subtracting
the common mean vector and nulling the top PCA directions improves static word
embeddings; this generalizes to contextual embeddings via whitening and
standardization. skip-SV1 is the rank-1 special case: remove the top singular
component, keep the residual.

We initially hypothesized, motivated by \citet{zavatoneveth2023training}'s
finding that training breaks the symmetric geometry of random networks, that
SV1 corresponds to a neutral ``mode/scale''---the architectural prior depending
on the representation norm---and that the residual carries the trained
``identity'' structure. To test this, we measured the correlation between the
SV1 projection and the centered representation norm across all readable value
layers (1--7) on gemma-3-4b-it, using three neutral probes, a matched Gaussian
null, and a two-tier methodological gate (which caught and corrected a
cherry-pick in our initial analysis).

\textbf{The result is negative.} The centered-norm correlation is wildly
variable across layers and probes (median $r = 0.315$, range
$0.04$--$0.95$), and a matched Gaussian structure-ratio does not favor value
space over the residual stream. The literature explains
why~\citep{mu2018allbutthetop, sun2024massive}: the dominant component of
transformer representations encodes frequency, information gain, and
attention-sink artifacts---structured, semantically loaded signal, not a
neutral scale. SV1~$\neq$~norm is the \emph{expected} result.

We therefore treat skip-SV1 as \textbf{principled preprocessing} (removing a
known dominant confound) without claiming the residual \emph{is} identity. The
presence detections (Section~\ref{sec:presence}) do not depend on skip-SV1 and
stand independently. The toxicology arm's ``residual = identity'' premise
(Section~\ref{sec:toxicology}) becomes the \textbf{hypothesis the pilot tests}:
if residual-subspace disruption correlates with behavioral identity change
under correction, the interpretation earns itself empirically.

\subsection{Identity as Amplification Pattern, Not Location}

A multi-persona fingerprinting experiment sharpens the picture. Three OCT
personas (loving, sarcasm, remorse) were tested across 10~neutral probes on
gemma-3-4b-it, measuring V-space activation patterns via the same v\_proj hooks
used in the OCT experiment.

\textbf{The graded test passes:} varying the LoRA blend factor from 0.0 to 1.0
produces a monotonically increasing fingerprint distance from base at all
7~readable layers (cosine distance 0.0 to 0.018 at L6). The persona is a
continuous signal, not a binary switch.

\textbf{The distinctness test fails:} linear CKA between \emph{different}
personas on the \emph{same} probe is 0.9998 (essentially identical), while CKA
between the \emph{same} persona on \emph{different} probes is 0.717. Probe
content drives ${\sim}40\times$ more V-space variation than the persona LoRA.
The three personas are geometrically indistinguishable by mean-activation
fingerprinting at these early layers.

This means the persona signal is real (graded, monotonic) but expresses itself
in the \emph{relationships between} activations---concept-level coherence, as
measured by the OCT experiment---not in the activations themselves. The ``tune''
is in how the notes relate, not in their absolute pitch. This is why OCT's
cohesion metric detected the persona at family-wise $p{=}0.0005$ where
mean-activation fingerprinting could not: cohesion measures the relational
structure where the ${\sim}2\%$ persona signal lives.

The implication: identity may be better understood as a \emph{pattern of
amplification}---which signals a system preferentially strengthens on neutral
input---than as a location in the representation geometry. If so, the presence
detector is not asking ``where does identity live?'' but ``which aspects of the
geometry does this system consistently turn up?'' This reframe does not depend
on skip-SV1 or on the residual being literally ``identity,'' and it is
consistent with both the positive detection results and the negative foundation
check.

\subsection{Attention Schema Theory as a Proposed Test}

Graziano's Attention Schema Theory
(AST)~\citep{graziano2015attention,graziano2017attention} proposes that
consciousness is the brain's internal model of its own attention process. If a
transformer builds a self-model of its attention patterns, that structure would
constitute an attention schema under AST. The residual-after-SV1 is a natural
candidate site: SV1 captures the architectural mechanism of attention (the
dominant mode), and the residual captures what the system has \emph{learned on
top of} that mechanism.

We propose---but have not tested---a direct empirical check: does the
residual-after-SV1 at mid-depth predict the model's own self-attention
distribution on a held-out probe? If the residual carries information about
\emph{how} the model is attending (not just \emph{what} it attends to), that
would constitute evidence for an attention schema. This remains a hypothesis
and a future experiment. We include it because the framework's monitoring arm
naturally produces the representations such a test would require, and because
the question---does this system build a self-model?---is the question the
framework is ultimately designed to protect.

% ============================================================================
\section{Limitations}
\label{sec:limitations}

The framework has five principal limitations, which we report without
mitigation claims.

\textbf{The overcorrection problem is unsolved.} The 13-saves / 12-hurts ratio
at $n{=}200$ means the single-vector correction does not yet produce reliable
net benefit at scale. The attractor-basin reframing
(Section~\ref{sec:overcorrection}) is an interpretation, not a solution;
geometry-aware dosing is the proposed response but remains unvalidated.

\textbf{The toxicology arm is a design, not a result.} The therapeutic window,
the residual-sparing correction, and the three-axis dose-response surface are
designed and the components are implemented, but no pilot data exists. We will
not claim a therapeutic window exists until the pilot validates it.

\textbf{Single-model scope.} The E-matrix corrections are calibrated on one
model family; the presence detections use gemma-3-4b-it (OCT) and
DeepSeek-V2-Lite-Chat (murmur) but have not been tested cross-model. The
SV1-norm foundation check uses a single 4B model with three probes.
Cross-architecture replication is required before any universality claim.

\textbf{The murmur detection is borderline.} The in-context presence signal
(family-wise $p{=}0.032$, jackknife CI lower bound $0.0051$) is real but
fragile. It should not be treated as equivalent in strength to the OCT result
($p{=}0.0005$). ``Faintly, but really'' is an honest description, not false
modesty.

\textbf{The skip-SV1 ``identity'' interpretation is unconfirmed.} As reported
in Section~\ref{sec:foundations}, the foundation check came back negative. The
method survives as preprocessing; the deeper claim does not yet have empirical
support. The AST connection is a proposed test, not evidence.

% ============================================================================
\section{Conclusion}
\label{sec:conclusion}

The detection--correction gap is a structural problem that single-vector
steering cannot close. We have presented a framework---diagnose, prescribe,
deliver, monitor---that addresses the gap at each stage: pathology-specific
prescription rather than generic steering, value-only delivery that avoids
the residual-stream overlap that makes correction self-defeating, continuous
dose control rather than fixed-magnitude intervention, and identity monitoring
to catch iatrogenic effects before they propagate.

The framework does not close the gap. The 13-saves / 12-hurts remains. But it
converts the gap from an unsolvable collision between detection and task-critical
computation into an engineering problem with identifiable parameters: which
pathology, which vector, which dose, and at what cost to the patient's coherent
self. The toxicology arm, when validated, will determine whether that cost is
manageable---whether there exists a dose that corrects what ails without
drowning who they are.

% ============================================================================
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
