We present a lightweight, model-agnostic method for detecting confabulation
in language models at inference time, requiring no access to model weights,
no training data, and no modification to the model itself. Singular value
decomposition (SVD) of the key-value cache---a standard byproduct of
autoregressive generation available on any transformer backend---produces
spectral features that distinguish genuine reasoning from confabulation
with under 50ms overhead per response. We introduce Marchenko-Pastur (MP)
corrected features derived from random matrix theory that are approximately
dimension-invariant, eliminating the dominant token-count confound that
plagued prior KV-cache analysis. In a clean experiment across three
architectures (Qwen2.5-7B, Llama-3.1-8B, Mistral-7B-v0.3; 467 trials),
post-hoc behavioral classification of naturally-occurring hedged versus
confabulated responses yields leave-one-out AUROC of 0.707 (Qwen,
$p=0.001$) using MP features alone, with encoding-phase features at chance
(0.511, $p=0.294$), confirming that the signal is generation-specific. The
method works as a drop-in safety layer: any system serving a transformer
model can extract KV-cache geometry between generation and response
delivery, flag likely confabulation, and optionally trigger regeneration.
We further describe the Oracle Loop, an architecture that extends this
monitoring into a complete self-regulation pipeline with activation
steering and geometry-aware GRPO training rewards. We report honestly on
three rounds of adversarial red-teaming that identified and corrected
critical methodological flaws, including a row-permutation invariance bug
in spectral null testing. All code, data, and red-team logs are publicly
available.
