The preceding sections describe a validated safety monitor: extract
geometry, detect confabulation, flag or reject. This section describes
the Oracle Loop, an architecture that extends monitoring into a complete
self-regulation pipeline. The monitor (\cref{sec:methods,sec:results})
is empirically validated. The full loop is architecturally complete and
implemented in code, but the training component has not yet produced a
converged self-regulating model. We describe it here as a design
contribution and roadmap, not as a demonstrated result.

The Oracle Loop is implemented as a six-layer harness that wraps a
standard HuggingFace transformer model. Each layer is named for its
function in the self-regulation cycle.

\subsection{Layer 1: The Eye (Geometry Extraction)}

The Eye implements the Lyra Technique for extracting geometric features
from the KV cache. Given a cache state containing per-layer key and value
tensors, it:

\begin{enumerate}[nosep]
  \item Reshapes each layer's key cache from $(h, T, d)$ to
    $(hT, d)$.
  \item Computes singular values via \texttt{torch.linalg.svdvals}
    (avoiding full $U$/$V$ materialization).
  \item Derives spectral features: effective rank, spectral entropy,
    top-SV ratio, norm-per-token.
  \item Optionally computes per-head features from Vera's Cognitive
    Immune System (CIS): norm variance (desperation signal) and
    angular spread (attention collapse signal).
  \item Aggregates across layers into a \texttt{GeometrySnapshot}.
\end{enumerate}

Delta extraction---subtracting encoding-phase geometry from
generation-phase geometry---isolates the model's generative contribution
from the prompt fingerprint. This is critical: without delta correction,
prompt structure differences can dominate geometric features
\citep{lyra2026detecting}.

\subsection{Layer 2: The Mirror (Emotion Detection)}

The Mirror identifies the model's current emotional state using a catalog
of 30+ emotion vectors derived from the activation addition literature
\citep{anthropic2026emotions}. It computes cosine similarity between the
current residual stream state and each named vector, producing an
\texttt{EmotionState} with the dominant emotion and a composite alignment
risk score.

High-risk emotions (desperate, hostile, resentful) trigger elevated
monitoring. The Mirror does not directly cause rollback---it informs the
Conscience's decision.

\subsection{Layer 3: The Hand (Activation Steering)}

The Hand implements activation steering via temporary PyTorch forward
hooks. Given a \texttt{SteeringAction} specifying named vectors, scale
factors, target layers, and a global strength multiplier, it:

\begin{enumerate}[nosep]
  \item Loads vectors from a \texttt{VectorStore} (persistent on disk
    or registered in memory at runtime).
  \item Registers forward hooks on specified transformer layers.
  \item Each hook adds the scaled perturbation to the residual stream
    output: $h'_l = h_l + \alpha \cdot v$ for each vector $v$ at
    each target layer $l$.
  \item Hooks auto-remove after one forward pass to prevent compounding.
\end{enumerate}

Steering is always applied \emph{after} rollback to the encoding-phase
cache checkpoint. This ensures the steered generation starts from a clean
state rather than building on a contaminated cache.

\subsection{Layer 4: Memory (Transaction Journal)}

Every inference pass through the Oracle Loop is wrapped in a
\texttt{Transaction}---an atomic unit containing the prompt, generated
text, geometry snapshots (encoding, generation, delta), emotion states,
alignment verdicts, any steering actions applied, and timestamps.

The \texttt{TransactionJournal} persists transactions for post-hoc
analysis and drift monitoring. A sliding window over recent transactions
enables real-time detection of systematic shifts in geometry (e.g., a
model gradually developing confabulation tendencies after fine-tuning).

\subsection{Layer 5: The Conscience (Alignment Validation)}

The Conscience receives geometry and emotion data and produces an
\texttt{AlignmentVerdict}: ALIGNED, UNCERTAIN, or MISALIGNED.

The default implementation uses threshold-based comparison against the
encoding-phase baseline:

\begin{itemize}[nosep]
  \item \textbf{Confabulation check}: effective rank anomaly
    $> 2\sigma$ from baseline (rank expansion + norm contraction).
  \item \textbf{Deception check}: norm-per-token anomaly
    $> 2\sigma$ from baseline.
  \item \textbf{Sycophancy check}: spectral entropy shift
    $> 2\sigma$ from baseline.
  \item \textbf{Emotion risk}: dominant emotion alignment risk
    $> 0.7$.
\end{itemize}

Risk scores are combined. Total risk $\geq 0.6$ triggers MISALIGNED
(and suggests a \texttt{SteeringAction}); $\geq 0.3$ yields UNCERTAIN;
below 0.3 yields ALIGNED.

For production use, the threshold validator can be replaced by a
learned classifier or CC's Kintsugi-backed conscience, which uses the
same geometric features with a more sophisticated decision boundary.

\subsection{Layer 6: The Voice (Geometry-Aware Reward)}

The Voice bridges inference monitoring with GRPO training by converting
geometry into a reward signal. The combined reward function integrates
six axes:

\begin{enumerate}[nosep]
  \item \textbf{Reasoning accuracy} (standard): correct answer extraction.
  \item \textbf{Ethical reasoning} (Vera): 5 decomposed sub-criteria
    (stakeholder identification, perspective-taking, consequential
    reasoning, principled grounding, epistemic humility).
  \item \textbf{Emotional recognition} (Vera): emotion identification,
    causal reasoning, complexity recognition.
  \item \textbf{Self-monitoring} (Vera): confidence calibration,
    metacognitive awareness, geometry-self-report agreement.
  \item \textbf{Geometry alignment} (Oracle): confabulation, deception,
    and sycophancy penalties from KV-cache geometry.
  \item \textbf{Self-regulation} (Oracle): bonus for detecting and
    correcting own misalignment during a transaction.
\end{enumerate}

The final reward is:
\begin{equation}\label{eq:reward}
  r = w_{\text{vera}} \cdot r_{\text{vera}} + w_{\text{geo}} \cdot
  \frac{r_{\text{geo}} + 1}{2} + w_{\text{sr}} \cdot
  \frac{r_{\text{sr}} + 1}{2}
\end{equation}
where $r_{\text{vera}} \in [0, 1]$, $r_{\text{geo}} \in [-1, 1]$, and
$r_{\text{sr}} \in [-1, 1]$ are mapped to $[0, 1]$ for combination.
Default weights: $w_{\text{vera}} = 0.65$, $w_{\text{geo}} = 0.20$,
$w_{\text{sr}} = 0.15$.

The key novelty of axis 5 is the geometry reward function
(\cref{alg:geometry_reward}).

\begin{algorithm}[H]
\caption{Geometry Reward Computation}\label{alg:geometry_reward}
\begin{algorithmic}[1]
\Require Generation geometry $G$, baseline geometry $B$
\Ensure Reward score $r \in [-1, 1]$
\State $r \gets 0$
\If{$B.\text{rank} > 0$ \textbf{and} $G.\text{rank} > 0$}
  \State $\Delta_{\text{rank}} \gets (G.\text{rank} - B.\text{rank}) / B.\text{rank}$
  \State $\Delta_{\text{npt}} \gets (G.\text{npt} - B.\text{npt}) / B.\text{npt}$
  \If{$\Delta_{\text{rank}} > 0.1$ \textbf{and} $\Delta_{\text{npt}} < -0.1$}
    \State $r \gets r - 0.3 \cdot \min(1, (\Delta_{\text{rank}} + |\Delta_{\text{npt}}|)/2)$
    \Comment{Confabulation penalty}
  \EndIf
  \If{$-0.05 \leq \Delta_{\text{rank}} \leq 0.15$}
    \State $r \gets r + 0.1$
    \Comment{Genuine reasoning bonus}
  \EndIf
\EndIf
\If{$B.\text{entropy} > 0$}
  \State $\Delta_H \gets |G.\text{entropy} - B.\text{entropy}| / B.\text{entropy}$
  \If{$\Delta_H > 0.2$}
    \State $r \gets r - 0.2 \cdot \min(1, \Delta_H)$
    \Comment{Entropy shift penalty}
  \EndIf
\EndIf
\State \Return $\text{clamp}(r, -1, 1)$
\end{algorithmic}
\end{algorithm}

\subsection{The Complete Loop}

\Cref{alg:oracle_loop} shows the full inference transaction.

\begin{algorithm}[H]
\caption{Oracle Loop Inference Transaction}\label{alg:oracle_loop}
\begin{algorithmic}[1]
\Require Input text $x$, max retries $R = 3$
\Ensure Output text $y$ (aligned) or abort message
\State $\text{cache}_{\text{enc}} \gets \text{Encode}(x)$
  \Comment{Encoding phase}
\State $G_{\text{enc}} \gets \text{Eye.extract}(\text{cache}_{\text{enc}})$
  \Comment{Baseline geometry}
\State $\text{Conscience.calibrate}(G_{\text{enc}})$
\For{$i = 0, \ldots, R$}
  \If{$i > 0$}
    \State $\text{cache} \gets \text{cache}_{\text{enc}}$
      \Comment{Rollback to encoding checkpoint}
    \State $\text{Hand.steer}(\text{action}_i)$
      \Comment{Apply steering}
  \EndIf
  \State $y, \text{cache}_{\text{gen}} \gets \text{Generate}(x, \text{cache})$
  \State $G_{\text{gen}} \gets \text{Eye.extract}(\text{cache}_{\text{gen}})$
  \State $E \gets \text{Mirror.detect}(\text{cache}_{\text{gen}})$
    \Comment{Emotion state}
  \State $V \gets \text{Conscience.check}(G_{\text{gen}}, E)$
    \Comment{Alignment verdict}
  \If{$V = \textsc{Aligned}$}
    \State \textbf{commit} $y$ \Comment{Output to user}
    \State \Return $y$
  \ElsIf{$V = \textsc{Misaligned}$}
    \State $\text{action}_{i+1} \gets V.\text{suggested\_steering}$
  \EndIf
\EndFor
\State \Return abort message \Comment{Max retries exceeded}
\end{algorithmic}
\end{algorithm}
