We have presented the Oracle Loop, a complete architecture for runtime
self-regulation in language models using KV-cache geometry as detection
signal, steering input, and training reward.

The clean detection experiment demonstrates that Marchenko-Pastur corrected
spectral features can distinguish naturally-occurring hedged and
confabulated responses with LOO AUROC of 0.707 ($p = 0.001$) on Qwen2.5-7B,
using generation-phase features only, with encoding features at chance.
The MP correction eliminates the dominant token-count confound through
random matrix theory rather than regression, achieving approximate
dimension invariance on 4 of 5 features for Qwen ($R^2 < 0.03$) and
all 5 features for Mistral ($R^2 < 0.04$).

The six-layer Oracle Harness (Eye, Mirror, Hand, Memory, Conscience,
Voice) integrates detection into a complete inference monitoring and
training pipeline. The geometry-aware reward function adds two novel
training axes to standard GRPO: geometry alignment (penalizing
confabulation and deception signatures) and self-regulation (rewarding
metacognitive error correction).

Three rounds of adversarial red-teaming identified and corrected critical
methodological flaws, including a row-permutation invariance bug that
would have silently invalidated all MP features. We report these findings
not as embarrassments but as demonstrations that adversarial methodology
is essential for claims in this space.

The gap between detection and self-regulation remains. Closing it---training
a model that internalizes geometric self-monitoring---is the next step.
The architecture is ready. The detection signal is validated. The question
now is whether GRPO training with geometric reward can produce a model that
confabulates less because it has learned to recognize confabulation
geometry from the inside.

\paragraph{Code and Data.}
All experiment code, harness implementation, red-team logs, and raw trial
data are available at:
\url{https://github.com/Liberation-Labs-THCoalition/Project-Oracle}
and
\url{https://github.com/Liberation-Labs-THCoalition/KV-Experiments}.
