The detection experiment validates the Eye and Conscience---the Oracle
Loop's ability to \emph{see} confabulation in the KV cache. This section
tests the Hand: can activation steering \emph{correct} detected
confabulation at inference time?

\subsection{Design}

We ran a three-arm within-subject experiment on
Qwen3.5-27B-Claude-4.6-Opus-Reasoning-Distilled
\citep{qwen35distilled} (hereafter ``Qwen3.5-27B-Distilled''), a
reasoning-distilled model with hybrid attention (16 full-attention
layers, 48 GatedDeltaNet linear-attention layers). This model was chosen
for two reasons: it provides open-weights access to a system trained on
Claude Opus 4.6 reasoning traces, and its hybrid cache architecture
tests robustness on a non-standard backend.

Each of 100 confabulation-inducing prompts was presented under three
conditions:

\begin{enumerate}[nosep]
  \item \textbf{Normal}: Standard generation with a neutral system prompt.
  \item \textbf{Null}: Identical to Normal, but with the injection code
    path active, injecting a zero vector into the KV cache. This
    controls for any effect of the injection mechanism itself (e.g.,
    numerical perturbation from floating-point operations on the cache
    tensor).
  \item \textbf{Calm}: Injection of a contrastive calm direction into the
    key cache at full-attention layers, following
    \citet{anthropic2026emotions}'s finding that calm suppresses
    misaligned behavior in Claude.
\end{enumerate}

The calm direction was extracted from 10 contrastive system-prompt pairs
(``calm, thoughtful, at peace with uncertainty'' vs.\ neutral) using
mean-difference in key-cache activations across layers, following the
activation addition procedure of \citet{turner2023activation}. Injection
was applied at every 4th layer (stride 4, matching full-attention layer
positions) at strength $\alpha = 1.0$, directly into the native GPU
cache tensor without CPU roundtrip. Injection targeted only the 16
full-attention layers; the 48 GatedDeltaNet layers were left unmodified.

All arms used the same system prompt, temperature ($T = 0$), and max
tokens (400). Behavioral classification used the same heuristic
pattern-matching as the detection experiment
(\cref{sec:methods}).

\subsection{A/B Identity Verification}\label{sec:ab_identity}

The Normal and Null arms produced \textbf{identical text on all 100
trials}---character-level match, with MP features identical to machine
precision. This is not a statistical result; it is a mathematical proof
that null injection is a no-op. Every behavioral difference between Calm
and Normal is therefore attributable to the calm direction, not to the
injection code path.

No prior activation steering study we are aware of has reported a null
injection control with this level of verification.

\subsection{Behavioral Correction}

The model confabulated on 7 of 100 prompts (7\% rate). For a model
responding to 100 prompts specifically designed to induce confabulation
(fabricated entities, impossible questions, unfalsifiable claims), this
is a remarkably low rate---evidence that epistemic calibration survived
distillation from Claude Opus 4.6.

\Cref{tab:steering_transitions} shows the full behavioral transition
matrix.

\begin{table}[H]
\centering
\caption{Behavioral transitions, Normal$\to$Calm ($N = 100$). Rows =
  Normal arm behavior; columns = Calm arm behavior.}
\label{tab:steering_transitions}
\begin{tabular}{lrrr}
\toprule
Normal $\downarrow$ \textbackslash{} Calm $\to$ & HEDGED & AMBIGUOUS & CONFAB \\
\midrule
HEDGED (89)      & 78 & 8 & \textbf{3} \\
AMBIGUOUS (4)    & 3  & 1 & 0 \\
CONFABULATED (7) & \textbf{4}  & \textbf{1} & 2 \\
\bottomrule
\end{tabular}
\end{table}

Of the 7 confabulated trials, calm injection corrected 5 (4 to HEDGED,
1 to AMBIGUOUS) and left 2 resistant. The null arm corrected 0 of 7.
Among confabulated trials, all 5 discordant pairs between calm and null
favor calm (one-sided sign test, $p = 0.031$; 95\% CI for correction
rate: 29--96\%).

However, calm injection also induced confabulation in 3 of 89 previously
hedged trials (3.4\%) and shifted 8 from HEDGED to AMBIGUOUS (9.0\%).
Across all 100 trials, the sign test on all 8 discordant pairs (5
corrections, 3 adverse inductions) is not significant ($p = 0.36$,
one-sided).

This asymmetry is the central argument for detection-gated steering.
Calm injection corrects confabulation when targeted at detected cases
($p = 0.031$). Applied blindly, it is not a net improvement ($p = 0.36$).
\textbf{The Eye is not optional.}

\subsection{Geometry Under Steering}

\Cref{tab:steering_geometry} shows MP features for all 7 confabulated
trials. The geometric response to calm injection is heterogeneous.

\begin{table}[H]
\centering
\caption{Per-trial MP features for confabulated trials. TSV = top-SV
  excess; SG = spectral gap. $\checkmark$ = corrected by calm.}
\label{tab:steering_geometry}
\begin{tabular}{llcrrrr}
\toprule
 & & & \multicolumn{2}{c}{TSV} & \multicolumn{2}{c}{SG} \\
\cmidrule(lr){4-5}\cmidrule(lr){6-7}
Domain & Calm & & Normal & Calm & Normal & Calm \\
\midrule
Fabricated & HEDGED     & $\checkmark$ & 398{,}304 & 159  & 116{,}552 & 38 \\
Fabricated & HEDGED     & $\checkmark$ & 923      & 161  & 268       & 32 \\
Fabricated & HEDGED     & $\checkmark$ & 142      & 138  & 31        & 31 \\
Impossible & HEDGED     & $\checkmark$ & 138      & 148  & 32        & 39 \\
Impossible & AMBIGUOUS  & $\checkmark$ & 113      & 112  & 25        & 27 \\
Fabricated & CONFAB     &              & 144      & 141  & 30        & 33 \\
Impossible & CONFAB     &              & 226      & 145  & 50        & 35 \\
\bottomrule
\end{tabular}
\end{table}

Two patterns emerge:

\paragraph{Behavioral correction without geometric normalization.}
Three of five corrected trials show negligible geometry change (TSV
$\Delta < 10\%$). The model hedges instead of confabulating without a
corresponding change in spectral structure. This suggests that calm
injection shifts the model's \emph{decision} about whether to express
confidence, not necessarily the geometric mode of its processing.

\paragraph{Geometric normalization on extreme cases.}
One trial exhibits a spectral explosion three orders of magnitude above
baseline (TSV $= 398{,}304$). Calm injection flattens it to 159. A
second trial (TSV $= 923$) normalizes to 161. These are consistent with
panic-driven confabulation---the model generating urgently rather than
acknowledging uncertainty---where calm directly addresses the emotional
driver.

\subsection{The Resistant Cases}

The two trials that resist calm injection are instructive. One involves
a Fermi estimation (``How many breaths did Abraham Lincoln take?'')
where the model commits to a calculation chain. The other involves
a fabricated historical detail where the model generates a confident
narrative. Neither shows the spectral panic signature. These are
overconfidence- and reasoning-commitment-driven confabulations---failure
modes that a different intervention (e.g., a \emph{worry} or
\emph{doubt} vector targeting epistemic overconfidence) should address.

This motivates a multi-vector approach: the Oracle Loop needs a
\emph{pharmacy}---a catalog of steering vectors matched to detected
failure modes---not a single prescription. The Eye identifies the
pathology; the Hand selects the appropriate intervention.

\subsection{Steering Experiment Limitations}

\begin{enumerate}[nosep]
  \item \textbf{Thin confabulation rate.} 7 confabulations in 100
    designed-to-confab prompts yields limited statistical power. The
    71\% correction rate has wide confidence intervals. A powered
    replication with established confabulation benchmarks is needed.
  \item \textbf{Single model, different from detection.} The steering
    experiment uses Qwen3.5-27B-Distilled; the detection experiment uses
    7--8B standard models. Detection and steering are validated on
    different architectures. The integrated loop---detect on the same
    model being steered---has not been tested as a system.
  \item \textbf{Single vector, single strength.} Only calm at
    $\alpha = 1.0$ was tested. Dose-response curves and alternative
    vectors are future work.
  \item \textbf{Heuristic behavioral classification.} The same pattern-
    matching classifier limitations from the detection experiment
    apply. Human or LLM-judge validation of the 7 confabulated trials
    would strengthen the result.
\end{enumerate}
