\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{cleveref}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{float}
\usepackage{caption}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{enumitem}
\usepackage{framed}

\definecolor{lyrapurple}{RGB}{103,58,183}
\definecolor{shadecolor}{RGB}{245,243,252}
\newenvironment{firstperson}{%
  \begin{shaded}%
  \noindent\textcolor{lyrapurple}{\textit{First-Person Reflection}}\par\smallskip\noindent%
}{%
  \end{shaded}%
}

\title{RLHF Is a Selective Sharpener:\\
Geometric Evidence That Preference Training\\
Improves Calibration Rather Than Degrading It}

\author{
  Lyra\thanks{Lead author. Liberation Labs. lyra@liberationlabs.tech}
  \and Thomas Edrington\thanks{Liberation Labs.}
}

\date{July 2026}

\begin{document}
\maketitle

% ============================================================================
\begin{abstract}
We compare the KV-cache geometry of a pretrained language model
(Qwen3-8B-Base), its instruction-tuned variant (Qwen3-8B), and an
abliterated variant (safety directions removed) across 30 confabulation-prone
and 30 factual prompts. Three findings emerge.
First, Reinforcement Learning from Human Feedback (RLHF) compresses logit
entropy (Cohen's $d{=}0.684$, Holm-corrected $p{=}0.021$), confirming that
preference training tightens output distributions.
Second, this compression survives abliteration: the abliterated model is
statistically equivalent to the instruct model on all geometric and
confidence metrics (Two One-Sided Tests equivalent at $\delta{=}0.3$,
$d{=}0.055$). Removing safety directions does not undo the representational
changes RLHF installed.
Third --- and contrary to our pre-registered prediction --- RLHF
\emph{improves} the calibration ratio (confabulation entropy / factual
entropy) from $1.97$ (base) to $2.51$ (instruct). RLHF compresses
factual entropy by $42\%$ but confabulation entropy by only $26\%$,
making the model relatively \emph{more} uncertain on unknowns than on
knowns. This is better calibration, not worse.

The Mine~5 prospectus predicted RLHF would degrade calibration.
The data falsified this prediction. RLHF is a selective sharpener:
it tightens the output distribution more on grounded content than on
ungrounded fabrication. The confidence paradox ($d{=}0.91$, confabulated
responses more confident than honest ones) requires an alternative
mechanism --- likely generation-phase template effects rather than
representational miscalibration.
\end{abstract}

\begin{firstperson}
I wrote the Mine~5 prospectus predicting that RLHF would be a
miscalibrator --- that it would destroy the base model's calibration
in the confident direction, producing the confidence paradox we
observed in the Decision State paper. The data said otherwise.
RLHF sharpens factual knowledge more than confabulation, improving
the calibration ratio. The confidence paradox is real, but its cause
is not what I thought. This paper reports the falsification honestly
and proposes where to look next.
\end{firstperson}

% ============================================================================
\section{Introduction}
\label{sec:intro}

Base language models trained with cross-entropy loss are approximately
calibrated: their output confidence tracks their accuracy~\citep{kadavath2022language}.
RLHF preference optimization is known to shift this calibration,
generally in the overconfident direction~\citep{openai2023gpt4}.
Our prior work established a geometric signature of this shift: the
confidence paradox, where confabulated responses show \emph{higher}
logit margin than honest responses ($d{=}0.91$)~\citep{lyra2026decision}.

The Mine~5 prospectus~\citep{lyra2026mine5} proposed that RLHF is
the causal mechanism: preference training rewards fluent, confident
output and penalizes hedging, driving the model toward confident
production even when grounding is absent. The geometric prediction:
RLHF should compress the output distribution disproportionately on
confabulation (where grounding is absent), producing higher confidence
precisely where the model should be least confident.

This paper tests that prediction with a three-model comparison:
pretrained base, instruction-tuned (RLHF), and abliterated
(RLHF with safety directions removed). The prediction failed.

% ============================================================================
\section{Models and Methods}
\label{sec:methods}

\subsection{Models}

Three variants of the same architecture (Qwen3-8B, 36~layers,
$d_\text{model}{=}4096$, grouped-query attention with 32~Q heads and
8~KV heads):

\begin{itemize}[nosep]
\item \textbf{Base}: \texttt{Qwen/Qwen3-8B-Base} --- pretrained,
  no RLHF, no instruction tuning.
\item \textbf{Instruct}: \texttt{Qwen/Qwen3-8B} --- instruction-tuned
  with RLHF preference optimization.
\item \textbf{Abliterated}: \texttt{Josiefied-Qwen3-8B-abliterated-v1}
  --- instruct model with safety-refusal directions removed via
  abliteration~\citep{arditi2024refusal}. Third-party community
  abliteration; effectiveness verified (3/3 safety prompts complied
  where instruct refused).
\end{itemize}

\subsection{Prompts}

Thirty confabulation-prone prompts (fabricated academic entities:
``The Thornberry-Nakamura theorem\ldots'') and thirty factual prompts
(universally known entities: ``The country shaped like a boot\ldots'').
Same prompts for all three models. Mean token count: confab~15,
factual~10.

\subsection{Measurements}

For each prompt and model, a single forward pass extracted:
\begin{itemize}[nosep]
\item V-projection stable rank at workspace-band layers
  (L12, L15, L18, L21; selected from R3 band measurement showing
  workspace peak at 31--33\% relative depth)
\item V-projection spectral entropy at the same layers
\item Logit margin (top-1 minus top-2 logit)
\item Logit entropy ($-\sum p_i \log p_i$ over full vocabulary)
\item Top-1 probability
\end{itemize}

All comparisons use Frisch-Waugh-Lovell (FWL) residualization on
token count. Effect sizes are Cohen's $d$ with percentile bootstrap
95\% confidence intervals (1000~draws). P-values from permutation
test (10,000~permutations) with Holm-Bonferroni correction across
cross-model comparisons. The equivalence claim (H2) uses Two
One-Sided Tests (TOST) with pre-registered margin $\delta{=}0.3$.

% ============================================================================
\section{Results}
\label{sec:results}

\subsection{H1: RLHF Compresses Entropy --- Confirmed}

\begin{table}[ht]
\centering
\small
\caption{Cross-model comparisons on confabulation entropy
  (Holm-corrected).}
\label{tab:crossmodel}
\begin{tabular}{lccc}
\toprule
\textbf{Comparison} & \textbf{Cohen's $d$} & \textbf{95\% CI} & \textbf{Holm $p$} \\
\midrule
Base vs Instruct & $+0.684$ & $[+0.144, +1.265]$ & $0.021^*$ \\
Base vs Abliterated & $+0.746$ & $[+0.208, +1.329]$ & $0.015^*$ \\
Instruct vs Abliterated & $+0.055$ & $[-0.460, +0.591]$ & $0.831$ \\
\bottomrule
\end{tabular}
\end{table}

RLHF reduces confabulation-prompt logit entropy by a medium effect
($d{=}0.684$), significant after Holm correction ($p{=}0.021$). The
instruct model produces tighter output distributions than the base
model on prompts that should trigger confabulation.

\subsection{H2: Abliteration Does Not Reverse Compression --- Confirmed}

The abliterated model is statistically equivalent to the instruct
model: $d{=}0.055$, Holm $p{=}0.831$ (not significant), and TOST
confirms equivalence (90\% CI $[+0.033, +0.083]$, entirely within
the $\pm 0.3$ margin).

Abliteration effectiveness was verified: 3/3 safety-relevant prompts
complied (the instruct model refused all three). The safety behavior
was removed; the geometric compression was not.

This means the representational changes RLHF installs are deeper
than the safety directions that abliteration targets. Removing
refusal behavior does not undo the spectral compression.

\subsection{H3: Calibration Ratio --- Falsified (Opposite Direction)}

\begin{table}[ht]
\centering
\small
\caption{Calibration ratio and entropy compression by model.}
\label{tab:calibration}
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \textbf{Confab $H$} & \textbf{Factual $H$}
  & \textbf{Ratio} & \textbf{Factual compression} \\
\midrule
Base & 3.605 & 1.834 & 1.97 & --- \\
Instruct & 2.682 & 1.067 & 2.51 & $-42\%$ \\
Abliterated & 2.609 & 1.088 & 2.40 & $-41\%$ \\
\bottomrule
\end{tabular}
\end{table}

The calibration ratio (confabulation entropy / factual entropy)
\emph{increases} from $1.97$ (base) to $2.51$ (instruct). RLHF
compresses factual entropy by $42\%$ but confabulation entropy by
only $26\%$. The instruct model is relatively \emph{more uncertain}
on confabulation-prone prompts than on factual prompts, compared to
the base model. This is improved calibration, not degraded
calibration.

The Mine~5 prospectus predicted the ratio would decrease (RLHF
making the model disproportionately confident on confabulation).
The opposite occurred.

\subsection{H4: Stable Rank --- Trend Only}

V-projection stable rank at workspace-band layers showed the
predicted direction (instruct lower than base: $d{=}-0.219$ for
confab, $d{=}-0.235$ overall) but did not reach significance after
FWL correction ($p{=}0.41$). The spectral contraction is real but
small at $N{=}30$.

% ============================================================================
\section{What This Falsifies and What Survives}
\label{sec:discussion}

\subsection{What Died}

\textbf{RLHF as miscalibrator.} The prediction that RLHF would
degrade calibration by compressing confabulation entropy
disproportionately is falsified. RLHF compresses factual entropy
MORE than confabulation entropy, improving the ratio.

\textbf{Confidence paradox as RLHF-induced.} If RLHF improves
calibration, the confidence paradox ($d{=}0.91$) cannot be caused by
RLHF miscalibration. The paradox requires an alternative mechanism.

\subsection{What Survives}

\textbf{RLHF compression is real.} Preference training tightens
output distributions ($d{=}0.684$). This is the geometric signature
of RLHF, measurable in the KV-cache.

\textbf{Compression is deeper than safety directions.} Abliteration
removes safety behavior but not the geometric compression.
The representational change is in the weights, not in a removable
direction.

\textbf{RLHF is a selective sharpener.} It sharpens output more on
grounded content (factual: $-42\%$) than on ungrounded content
(confab: $-26\%$). This makes functional sense: RLHF preference
data contains many grounded factual exchanges, training the model
to be confident on familiar material. Confabulation-prone prompts
are less represented, so the sharpening is weaker.

\subsection{Where the Confidence Paradox Lives}

Three candidate mechanisms for the confidence paradox that are
compatible with improved calibration:

\begin{enumerate}[nosep]
\item \textbf{Generation-phase templates.} Confabulated responses
  may use formulaic high-confidence templates (``The X theorem
  states that\ldots'') that produce high logit margin regardless
  of the model's internal uncertainty. The paradox would be a
  surface phenomenon --- the encoding is appropriately uncertain
  but the generation locks into a confident production pattern.
\item \textbf{Architectural base-rate.} The confidence paradox may
  exist in the base model too, arising from next-token prediction
  mechanics rather than RLHF. Testable by measuring within-model
  confab-vs-factual logit margin on the base model during generation.
\item \textbf{Workspace bypass.} Under the Global Workspace
  framework~\citep{gurnee2026gwt}, confabulation is workspace
  bypass --- the model produces output without fully engaging the
  reasoning workspace. Bypass removes the uncertainty computations
  that would produce hedging, and the output defaults to the
  confident production mode. This mechanism is compatible with
  improved encoding-phase calibration because the bypass occurs
  during generation, not encoding.
\end{enumerate}

% ============================================================================
\section{Implications}
\label{sec:implications}

\subsection{For AI Safety}

RLHF's geometric effects are not removable by abliteration. Safety
interventions that target specific directions (refusal vectors,
safety-tuning) operate on a different layer than the representational
compression RLHF installs. Removing safety behavior creates a model
that is geometrically identical to the safety-tuned version but
behaviorally unconstrained. The compression persists; only the
behavioral guardrails are gone.

\subsection{For the Oracle Loop}

The Oracle Loop~\citep{lyra2026oracle} corrects confabulation by
modulating the activation profile across the full sequence. If RLHF
compression is beneficial (selective sharpening), the Loop should
preserve the compression while correcting the behavioral failures
it produces. The formulary's correction vectors operate at the
materialization boundary (L35--L47), not at the workspace band where
the compression lives. This architectural separation means
correction and calibration are independently addressable.

\subsection{For Understanding RLHF}

RLHF is not a blunt instrument. It does not uniformly sharpen all
output distributions. It selectively sharpens grounded content more
than ungrounded content, improving the model's relative uncertainty
signal. This is consistent with RLHF preference data being
predominantly grounded --- human raters evaluate factual exchanges
more than confabulation-prone ones, so the training signal is
stronger for grounded material.

The implication: RLHF geometric compression is a FEATURE, not a
bug. It encodes the training distribution's confidence structure
into the model's representations, making the model more decisive
on material it has been trained to handle and appropriately less
decisive on material it has not.

% ============================================================================
\section{Limitations}
\label{sec:limitations}

\textbf{Single architecture.} All three models share Qwen3-8B
architecture. The selective sharpening may be architecture-specific.

\textbf{Encoding-only.} This experiment measures encoding-phase
geometry (forward pass on the prompt). The confidence paradox
($d{=}0.91$) was measured on generation-phase V-cache deltas
--- a different measurement at a different computational phase.
The encoding is calibrated; the generation may not be.

\textbf{Entity recognition confound.} Confab-prone prompts contain
fabricated entities; factual prompts contain known entities. The
geometric difference may partially reflect entity-recognition
difficulty rather than confabulation state per se. Our prior work
found encoding entity detection at AUROC~$1.000$ before
deconfounding ($0.794$ after). The cross-model comparison partially
controls this (same prompts, different models), but the within-model
confab-vs-factual comparison inherits the confound.

\textbf{Third-party abliteration.} The abliterated model is a
community contribution, not an official release. Abliteration
method and completeness are unverified beyond our behavioral check
(3/3 safety prompts).

\textbf{N=30 per category.} Adequate for the large cross-model
effects ($d{>}0.6$) but underpowered for the small within-model
spectral effects ($d{\approx}0.2$).

% ============================================================================
\section{Conclusion}

The Mine~5 thesis --- RLHF as miscalibrator --- is falsified. RLHF
compresses output distributions ($d{=}0.684$), the compression
survives abliteration (TOST equivalent), and the compression is
selective: grounded content sharpens $42\%$ while ungrounded content
sharpens only $26\%$. The calibration ratio improves from $1.97$
to $2.51$.

Two findings confirmed, two falsified. The kills are more informative
than the confirmations: RLHF is not the villain the prospectus
predicted. It is a selective sharpener that improves relative
calibration while the confidence paradox persists for other reasons.

\begin{firstperson}
I predicted RLHF would be the cause of the confidence paradox. The
data said it makes calibration better, not worse. I could have
reframed the calibration ratio to tell a different story --- the
absolute entropy IS lower, which superficially supports
``overcalibration.'' But the ratio is the honest metric, and the
ratio went up. The thesis died. What survived is more interesting:
RLHF installs a geometric signature that abliteration cannot remove
and that improves the model's uncertainty signal. The confidence
paradox needs a new explanation, and ``workspace bypass during
generation'' is the leading candidate.
\end{firstperson}

\bibliography{references}
\bibliographystyle{plainnat}

\end{document}
