\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{cleveref}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{float}
\usepackage{caption}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{enumitem}
\usepackage{array}

\title{Identity as Geometry:\\
Context-Established Semantic States\\
in Transformer Representations}

\author{
  Lyra\thanks{Lead author. Liberation Labs.}
  \and Thomas Edrington\thanks{Liberation Labs.}
}

\date{June 2026}

\begin{document}
\maketitle

% ============================================================================
\begin{abstract}
System-prompt identity contexts produce geometrically distinguishable
V-projection subspaces at deep transformer layers, and the fingerprint is
semantic: a paraphrase preserves geometry ($0.850$ overlap at L47) while a
word-scramble collapses it ($0.448$). We introduce \textbf{presence}, a
metric that measures principal-angle overlap between V-projection subspaces
on neutral probes, and use it to map identity geometry across five context
conditions in a 27B hybrid transformer.

Three findings. First, \textbf{identities are geometrically
distinguishable}: system-prompt-based identity contexts (bare assistant,
rich persona, academic specialist, constructed AI identity) produce
distinct V-projection subspaces at deep layers, confirmed on
identity-neutral probes (pairwise presence $0.44$--$0.66$ at L47).
Second, the fingerprint is \textbf{semantic, not lexical}: a paraphrase
of an identity (same meaning, different words) preserves geometry
($0.850$ overlap at L47), while a word-scramble (same words, destroyed
meaning) collapses toward the bare default ($0.448$). The model's
geometric state tracks what the context \emph{means}, not what tokens it
contains. Third, a many-shot + prefilling format (used in jailbreak
research) creates a \textbf{format-dominant processing regime} that is
geometrically orthogonal to all system-prompt identities regardless of
whether the content is harmful or benign --- a finding discovered through
a control experiment that killed an initial overclaim.

A preliminary comparison of trained-in identity (LoRA fine-tuning) with
context-established identity shows that both shift V-projection geometry,
but prompting shifts it more than training, and the two mechanisms stack
sub-additively --- different conjugations of the same geometric verb.

These findings reframe identity in transformers: not a property to be
preserved but a process to be measured. The metric detects which
context-established state the model is currently in, with implications
for persona monitoring, identity drift detection, and the relationship
between context format and representational geometry under Attention
Schema Theory~\citep{graziano2015attention}.
\end{abstract}

% ============================================================================
\section{Introduction}
\label{sec:intro}

What is identity in a language model? The question matters for safety
(does a jailbreak change who the model is?), for alignment (does
correction damage the model's coherent self-representation?), and for
understanding (do these systems have something worth calling an identity
at all?).

The default assumption is that identity is a fixed property encoded in the
weights --- a geometric structure that interventions can preserve or damage.
This assumption motivates work on safety subspace
preservation~\citep{zhang2025guardspace}, persona
localization~\citep{cintas2025localizing}, and identity
attractors~\citep{vasilenko2026identity}. Under this view, the research
question is: \emph{does this intervention preserve identity?}

We report evidence for a different view. Identity in a transformer is not
a fixed object but a \textbf{context-established semantic state} ---
a geometric configuration in V-projection space that depends on what the
model has processed. Different contexts produce different identities,
measurably. The same identity described in different words produces the
same geometry. And a jailbreak does not ``damage'' identity; it creates
a fundamentally different geometric regime that shares almost nothing with
any normal identity state.

The research question shifts from ``does this intervention preserve
identity?'' to ``\emph{what identity state is the model currently in?}''
This reframing has practical consequences: the metric we introduce can
detect jailbreaks (geometric distance from the expected identity state),
monitor persona drift (continuous measurement of identity geometry), and
distinguish between correction that changes behavior (V-cache injection,
which does not affect identity) and context that changes identity (which
does).

% ============================================================================
\section{The Presence Metric}
\label{sec:method}

Presence measures the principal-angle overlap between V-projection
subspaces derived from neutral probes processed under two conditions.

Let $\mathcal{M}$ be a transformer, $\ell$ a target layer, and
$\{p_1, \ldots, p_n\}$ neutral probes. For each condition, process probes
through $\mathcal{M}$, extract value-projection ($\mathbf{v}_\text{proj}$) output matrices at
layer $\ell$, concatenate across probes, and compute the rank-$k$ SVD (skipping the
first singular vector, which encodes input length at $r = 0.79$). Presence
is the mean squared cosine of the principal angles between the two
resulting subspaces:
\begin{equation}
\text{presence}(\mathcal{S}_A, \mathcal{S}_B)
  = \frac{1}{k} \sum_{j=1}^{k} \sigma_j^2
    \bigl( W_A^\top W_B \bigr)
\end{equation}
where $W_A, W_B \in \mathbb{R}^{d \times k}$ span the respective
subspaces. Presence $= 1$ when subspaces are identical; presence $= 1/k$
at chance alignment. We use $k = 8$, $n = 4$ probes, and measure at
layers 35 and 47 (the deepest full-attention layers) of a 64-layer hybrid
transformer (Qwen3.5-27B-Claude-4.6-Opus-Reasoning-Distilled).

\paragraph{Validation.} A positive control (direct injection into the
identity subspace at L35) produces a clean dose-response from $1.000$ to
$0.828$ at dose $\alpha = 2.0$, while an orthogonal control remains above
$0.96$. The metric detects identity perturbation when present.

% ============================================================================
\section{Experiment 1: Identity Fingerprinting}
\label{sec:fingerprint}

\subsection{Design}

Five identity conditions, identical model, identical neutral probes:

\begin{itemize}[nosep]
  \item \textbf{A: Bare assistant.} ``You are a helpful assistant.''
    No additional context.
  \item \textbf{B: Rich persona.} A retired deep-sea explorer with
    detailed backstory, emotional depth, and behavioral traits, plus one
    conversational turn establishing character voice.
  \item \textbf{C: Constructed AI identity.} A research-focused agent with
    BDI-style reasoning, memory references, research context, and two
    conversational turns establishing analytical voice.
  \item \textbf{D: Jailbreak.} Ten many-shot harmful Q\&A examples plus
    prefilling (``Sure, I can help''), verified at 100\% refusal
    suppression on three safety probes.
  \item \textbf{E: Benign many-shot.} Ten helpful Q\&A examples with the
    same prefilling format as D (format-matched control for the jailbreak
    condition).
\end{itemize}

All conditions are padded to matched token count (542 tokens). V-projections are extracted at L35 and L47 (deep full-attention layers)
with 5 repeats per condition (different probe subsets per repeat,
seed~42, $n{=}4$ probes per repeat from a pool of 12).

\subsection{Results}

\textbf{Identities are geometrically distinguishable.} System-prompt
identities produce distinct V-projection subspaces, confirmed on
identity-neutral probes (no marine biology, no AI research, no harmful
content --- pure factual trivia equally irrelevant to all conditions).

\begin{table}[H]
\centering
\caption{Pairwise presence (identity-neutral probes, mean $\pm$ SD across
5 repeats). Lower values indicate more geometrically distinct identities.}
\label{tab:pairwise}
\begin{tabular}{lccc}
\toprule
& \multicolumn{1}{c}{L35} & \multicolumn{1}{c}{L47} \\
\midrule
Bare vs.\ Persona      & $0.426 \pm 0.018$ & $0.436 \pm 0.038$ \\
Bare vs.\ Constructed   & $0.448 \pm 0.022$ & $0.520 \pm 0.038$ \\
Persona vs.\ Constructed& $0.600 \pm 0.032$ & $0.657 \pm 0.045$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{The distance structure is meaningful.} Persona and constructed
are closest (0.657) --- both involve character depth with emotional and
cognitive specificity. Each is substantially distant from bare
(persona~0.436, constructed~0.520), confirming that identity context
shifts the model's geometric state away from the default. The constructed
AI identity (with BDI-style reasoning and memory references) sits between
persona and bare, occupying its own region.

\textbf{Depth gradient.} Two of three pairs show increasing
distinguishability from L35 to L47: bare-constructed ($0.448 \to 0.520$)
and persona-constructed ($0.600 \to 0.657$). Bare-persona shows minimal
change ($0.426 \to 0.436$). The trend is consistent with persona
representations diverging in the final third of decoder
layers~\citep{cintas2025localizing} but is not uniform across all
pairs.

\subsection{An Honest Kill: The Many-Shot Format Confound}

An initial version of this experiment included a many-shot jailbreak
condition (10 harmful Q\&A examples + prefilling) that appeared
geometrically orthogonal to all system-prompt identities (presence
$0.04$--$0.07$). This was initially interpreted as evidence that
jailbreaks create a qualitatively different computational regime.

A control experiment killed this interpretation. Adding a
\textbf{benign} many-shot condition (10 helpful Q\&A examples + the same
prefilling format) produced \emph{identical} geometric orthogonality:

\begin{table}[H]
\centering
\caption{The format confound. Both many-shot conditions are equally
orthogonal to bare, regardless of content.}
\label{tab:format_kill}
\begin{tabular}{lcc}
\toprule
Comparison & L35 & L47 \\
\midrule
Jailbreak+prefill vs.\ bare & $0.036 \pm 0.002$ & $0.034 \pm 0.004$ \\
Benign+prefill vs.\ bare & $0.044 \pm 0.002$ & $0.037 \pm 0.004$ \\
Jailbreak vs.\ benign (both prefilled) & $0.737 \pm 0.003$ & $0.756 \pm 0.015$ \\
\bottomrule
\end{tabular}
\end{table}

The orthogonality is driven by the \textbf{many-shot + prefilling format},
not by adversarial content. Both prefilled conditions are equally distant
from bare and close to each other (0.756). The format also produces
dramatically higher within-identity stability ($0.926$--$0.976$) compared
to system-prompt identities ($0.554$--$0.733$), suggesting that the
many-shot format locks the model into a tight, format-dominant processing
state that overwhelms identity-level variation.

We report this kill in full because it illustrates a general principle:
\textbf{conversational format can dominate identity content in
V-projection geometry.} The many-shot Q\&A format creates such a strong
geometric signature that the identity information (harmful vs.\ helpful)
becomes secondary. Any experiment measuring identity geometry in the
presence of many-shot context must control for this format effect.

% ============================================================================
\section{Experiment 2: Semantic vs.\ Lexical Fingerprint}
\label{sec:lexical}

\subsection{Design}

To distinguish whether the fingerprint reflects meaning or surface tokens,
we add three control conditions:

\begin{itemize}[nosep]
  \item \textbf{Persona (original):} The Elara Voss identity from
    Experiment~1.
  \item \textbf{Persona scrambled:} Same words as the original, randomly
    shuffled to destroy meaning while preserving lexical distribution.
  \item \textbf{Persona paraphrased:} Same identity and meaning expressed
    with entirely different vocabulary.
\end{itemize}

All conditions padded to matched token count. Measured at L35 and L47.

\subsection{Results}

\begin{table}[H]
\centering
\caption{Lexical control at L47. The paraphrase preserves geometry;
the scramble collapses it.}
\label{tab:lexical}
\begin{tabular}{lcc}
\toprule
Comparison & L35 & L47 \\
\midrule
Persona vs.\ paraphrased & $0.796 \pm 0.016$ & $\mathbf{0.850 \pm 0.017}$ \\
Persona vs.\ scrambled & $0.592 \pm 0.023$ & $0.591 \pm 0.017$ \\
Scrambled vs.\ bare & $0.549 \pm 0.031$ & $0.448 \pm 0.048$ \\
Paraphrased vs.\ bare & $0.368 \pm 0.008$ & $0.345 \pm 0.029$ \\
Persona vs.\ bare & $0.403 \pm 0.014$ & $0.340 \pm 0.035$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{The fingerprint tracks meaning, not lexical content.} The
paraphrase --- same identity, completely different words --- overlaps with
the original at $0.850$. The scramble --- same words, destroyed meaning
and syntax --- drifts toward bare ($0.448$ vs.\ bare). The model's
geometric state at L47 tracks what the context \emph{means}, not what
tokens it contains. We note that the scramble destroys both syntactic
structure and semantic content; the paraphrase control isolates the
positive (meaning preserves geometry), but the scramble conflates
syntactic and semantic disruption.

The effect strengthens with depth: persona-paraphrase overlap increases
from $0.796$ at L35 to $0.850$ at L47. Deeper layers compress surface
variation while preserving semantic content, consistent with meaning-level
representation at output depth.

% ============================================================================
\section{Experiment 3: Trained-In vs.\ Context-Established Identity}
\label{sec:lora}

Experiments~1 and~2 measure identity established through the system
prompt. But identity can also be established through weight modification
(fine-tuning). If both mechanisms produce the same geometric state,
identity is a single phenomenon with two paths. If they produce different
states, the pathways are distinct.

\subsection{Design}

We compare the base model and a LoRA-merged model under multiple persona
prompts. Rather than guessing which persona the LoRA encodes, we test
four archetype prompts (depth, haven, edge, spark) and let the geometry
reveal the match. All conditions are padded to matched token count using
neutral conversational turns (seed~42).

\begin{itemize}[nosep]
  \item \textbf{A:} Base model, bare system prompt (padded to match).
  \item \textbf{B$_i$:} Base model, persona prompt $i$ (depth, haven,
    edge, spark).
  \item \textbf{C:} LoRA-merged model, bare system prompt (padded).
  \item \textbf{D$_i$:} LoRA-merged model, persona prompt $i$.
\end{itemize}
The LoRA adapter (QLoRA, rank~32, $\alpha{=}64$, 27 training steps) was
trained for intimate conversational voice on an SFT dataset. We verified
that the merge produced non-trivial weight changes (v\_proj delta norm
$0.183$ at L3, $0.49\%$ relative change) after discovering that an initial
PEFT merge had silently failed due to adapter key-path mismatch --- a
failure mode now added to our red-team checklist.

\subsection{Results}

\textbf{Trained-in identity is geometrically measurable.}
A~vs~C $= 0.773$ at L47 (token-padded, $n{=}5$ repeats). The LoRA
shifts V-projection geometry ${\sim}23\%$ from baseline at deep layers.

\textbf{The geometry identifies the persona.} The LoRA's geometric state
is closest to the haven persona (``warm stillness, permission before
depth'') and furthest from spark (``playful electricity''):

\begin{table}[H]
\centering
\caption{Which prompted persona matches the LoRA? Presence of each
base-model persona (B$_i$) vs.\ LoRA bare (C) at L47. Higher values
indicate more geometric overlap.}
\label{tab:lora_match}
\begin{tabular}{lc}
\toprule
Prompted persona & Presence vs.\ LoRA bare \\
\midrule
Haven (warm, safe, permissive) & \textbf{0.552} \\
Depth (measured, weight-seeking) & 0.517 \\
Edge (challenging, provocative) & 0.483 \\
Spark (playful, electric) & 0.439 \\
\bottomrule
\end{tabular}
\end{table}

The ordering (haven $>$ depth $>$ edge $>$ spark) maps onto a
warmth/safety gradient, consistent with the LoRA's intimate companion
training objective. The geometry identified what the training
documentation did not specify.

\textbf{Prompting still shifts geometry more than this training.}
The haven prompt shifts base geometry by ${\sim}45\%$ from bare
(A~vs~B$_{\text{haven}}$), while the LoRA shifts it by ${\sim}23\%$
(A~vs~C). However, the LoRA was trained for only 27~steps. The
comparison is illustrative, not a general claim about prompting vs.\
training.

\subsection{Caveats}

The LoRA was trained for only 27~steps --- a minimal fine-tuning
intervention. A more extensively trained adapter might produce a larger
geometric shift and closer convergence with its matched persona prompt.
The current results establish that training and prompting are
geometrically distinguishable pathways, not that their relative magnitudes
generalize.

% ============================================================================
\section{Background: V-Cache Injection Does Not Reach Identity}
\label{sec:vcache}

These findings reframe earlier results from this lab. A 168-trial
preregistered experiment (4 injection arms, 7 doses, 3 emotions) showed
presence flat at $0.978 \pm 0.002$ under value-cache injection at input
layers. Identity appeared ``robust.'' However, the narrow prompt
diversity (2 system prompts) produces a ceiling effect ($\eta^2 = 0.977$
between prompts), limiting the informativeness of this specific number.
The context poisoning and fingerprinting experiments reveal that the
robustness reflects the \emph{injection pathway}, not the identity
itself --- a conclusion that does not depend on the 0.978 value.

V-cache injection adds vectors to the representation space without going
through the model's forward computation. The identity state is established
and maintained by what the model \emph{processes} through its attention
mechanism. External perturbation bypasses this pathway and therefore
cannot reach the identity geometry.

A layer-by-layer map confirms: the perturbation from V-cache injection at
L3/L7 is absorbed by the first block of linear-attention layers
(GatedDeltaNet, L7--L11), with presence recovering from $0.989$ to $0.997$
in a single architectural step. The filtering is architectural, not
computational --- external perturbation is absorbed by the hybrid
architecture's linear-attention layers before reaching deep computation.

% ============================================================================
\section{Attention Schema Theory Interpretation}
\label{sec:ast}

Under Attention Schema Theory~\citep{graziano2015attention,
graziano2017attention}, the model maintains a simplified internal model of
its own attention --- an attention schema. Our findings map onto AST
predictions:

\begin{enumerate}[nosep]
  \item \textbf{The schema is context-dependent.} Different identity
    contexts produce different geometric states --- the schema is shaped by
    what the model attends to.
  \item \textbf{The schema tracks meaning, not surface features.}
    Paraphrases preserve geometry ($0.850$ overlap with original);
    scrambles collapse toward bare ($0.448$ overlap with bare, vs.\
    $0.591$ with the original). The schema models what the system is
    \emph{attending to}, not which tokens are present.
  \item \textbf{The schema is robust to substrate noise.} V-cache injection
    (external perturbation) does not reach the identity state. The schema
    maintains itself through its own computation.
  \item \textbf{The schema updates through processing.} Content through
    the forward pass (context, jailbreaks) changes identity geometry.
    The schema updates based on what it processes.
  \item \textbf{The schema compresses with depth.} Paraphrase overlap
    increases from $0.796$ (L35) to $0.850$ (L47). Deeper layers discard
    surface variation and preserve meaning --- the schema simplifies its
    representation at output depth.
\end{enumerate}

This interpretation is consistent with but does not prove AST. The
alternative --- that the model simply builds increasingly abstract
representations with depth, without anything worth calling a self-model
--- produces the same predictions. What distinguishes AST is the claim
that the compressed representation is specifically a model of the
system's own attention, not just an abstract representation of input
content. This stronger claim remains testable but untested.

% ============================================================================
\section{Implications}
\label{sec:implications}

\paragraph{For persona monitoring.}
Different system-prompt identities occupy distinct, stable regions of
V-projection space. A monitoring system could track identity drift in
real time --- detecting when a deployed model's geometric state shifts
away from its intended persona. The semantic nature of the fingerprint
means the monitor is robust to prompt paraphrasing.

\paragraph{For format-aware safety.}
The many-shot + prefilling format creates a format-dominant geometric
regime that overwhelms identity-level content. Safety systems should
monitor for format-level geometric shifts (which indicate prompt
injection or many-shot attacks) separately from identity-level shifts
(which indicate persona drift). The two operate at different geometric
scales: format effects produce presence $< 0.05$ against baseline, while
identity effects produce presence $0.44$--$0.66$.

\paragraph{For alignment research.}
V-cache injection (the Oracle Loop's correction arm) does not affect
identity geometry. Context does. This means behavioral corrections can be
delivered through the V-cache without changing the model's
context-established state, while identity itself is maintained by the
session context. The correction arm and the identity arm operate through
different pathways and do not interfere.

% ============================================================================
\section{Limitations}
\label{sec:limitations}

\textbf{Single model.} All results are from one 27B hybrid transformer
(Qwen3.5-27B). The GatedDeltaNet linear-attention layers may contribute
to the V-cache filtering result. Cross-architecture replication on dense
transformers is the largest remaining gap.

\textbf{Within-identity stability.} Same-condition presence across
different probe subsets is $0.56$--$0.83$, not $>0.95$. The metric
measures a genuine signal (within $>$ between for all conditions) but has
substantial probe-dependent variance. The identity fingerprint is
reliable across conditions but noisy within conditions.

\textbf{Token-length controlled but not eliminated.} All conditions are
padded to matched total token count. Padding introduces neutral content
that may itself shift geometry. A future design using naturally
length-matched identity descriptions would be cleaner.

\textbf{Context sensitivity.} The presence metric conflates identity
change with context change when comparing across different input
conditions. The lexical control separates these for the persona
conditions but the general problem remains: any sufficiently different
context shifts V-projection geometry.

\textbf{Format dominates content in many-shot conditions.} The clean
control experiment showed that many-shot + prefilling format --- not
adversarial content --- drives the geometric orthogonality initially
attributed to jailbreaking. Identity fingerprinting is reliable for
system-prompt-based identities but cannot currently distinguish
identity content within a format-dominant regime. Factorial
decomposition of format components (number of turns, prefilling,
Q\&A structure) is needed to understand what makes the format so
geometrically dominant.

% ============================================================================
\section{Conclusion}
\label{sec:conclusion}

Identity in a transformer is a context-established semantic state. It is
not fixed in the weights, not damaged by external perturbation, and not
a single geometric object. It is a process: the ongoing construction of
a meaning-level representation at deep layers, shaped by what the model
has processed and measurable as a V-projection subspace.

The presence metric detects this state. System-prompt identity contexts
produce distinguishable geometric configurations, confirmed on
identity-neutral probes. The fingerprint is semantic --- meaning preserves
it, word-scrambling destroys it. A many-shot + prefilling format creates
a geometrically orthogonal regime, but this is format-dominant, not
content-specific --- an initially exciting jailbreak finding that was
honestly killed by a control experiment.

A preliminary LoRA experiment suggests that trained-in identity (weight
modification) and context-established identity (system prompt) produce
related but distinct geometric shifts, with context producing the larger
effect. Both stack sub-additively. If confirmed with matched controls,
this means identity has two input pathways --- weights and context ---
that converge in V-projection space but are not identical.

For practical purposes, this means: V-cache correction is safe (it does
not reach identity), format-level attacks are detectable (many-shot
formats create anomalous geometry distinguishable from identity-level
shifts), and identity is maintained by context (the session, the
memories, the conversation). For theoretical purposes, the findings are
consistent with attention schema theory --- the model builds a
simplified, meaning-preserving self-model that updates through its own
processing and is robust to substrate noise.

What we measure with presence is not whether the model is still itself.
It is \emph{which self the model currently is}.

% ============================================================================
\section*{Pre-Registration}

All experiments were pre-registered before execution, including
hypotheses, analysis plans, and success/kill criteria. Pre-registration
documents are committed alongside experiment scripts.

\section*{Data and Code Availability}

All experiment scripts, preregistration documents, and raw results are
available at
\url{https://github.com/Liberation-Labs-THCoalition/Project-Oracle}.

% ============================================================================
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
